# Introduction

`r if(identical(gitbook_on, FALSE)) knitr::asis_output(paste0("This report is available as a PDF and as an online interactive report at ", ngr::ngr_str_link_url(url_base = params$report_url, anchor_text = params$report_url), ". We recommend viewing online as the web-hosted HTML version contains more features and is more easily navigable. Please reference the website for the latest PDF from ", ngr::ngr_str_link_url(url_base = params$repo_url, url_resource = "raw/main/docs/fish_passage_peace_2024_reporting.pdf", anchor_text = "fish_passage_peace_2024_reporting.pdf"), "."))`

`r if(identical(gitbook_on, FALSE)){knitr::asis_output("<br>")}`

```{r proj-partners-text-into, results='asis'}

project_partners_text <- dplyr::case_when(params$project_region == "fraser" ~ "Since 2023, the Society for Ecosystem Restoration Northern British Columbia (SERNbc), with support from the Ministry of Transportation and Infrastructure, has been actively involved in planning, coordinating, and conducting fish passage restoration efforts within the Nechako River, Lower Chilako River, Morkhill River, Upper Fraser River, and François Lake watershed groups which are sub-basins of the Upper Fraser River watershed. ",
                                          params$project_region == "peace" ~ "The Society for Ecosystem Restoration in Northern BC (SERNbc) is working together with the McLeod Lake Indian Band, the Peace Region Fish and Wildlife Compensation Program (FWCP), the Provincial Fish Passage Technical Working Group (FPTWG), road/rail tenure holders and other stakeholders/partners to prioritize, plan and fund the restoration of fish passage at road crossing structure barriers within the FWCP Peace Region. ",
                                          params$project_region == "skeena" ~ "Since 2020, the Society for Ecosystem Restoration Northern British Columbia (SERNbc) has been actively involved in planning, coordinating, and conducting fish passage restoration efforts within the Bulkley River and Morice River watershed groups, which are sub-basins of the Skeena River watershed. In 2022, the study area was expanded to include the Zymoetz River watershed group and the Kispiox River watershed group, followed by an extension in 2023 to encompass sections of the Kitsumkalum River watershed group, particularly where Highway 16 intersects the watershed."
)

cat(project_partners_text)
```

<br>

The primary objective of this project is to identify and prioritize fish passage barriers within study areas, develop comprehensive restoration plans to address these barriers, and foster momentum for broader ecosystem restoration initiatives. While the primary focus is on fish passage, this work also serves as a lens through which to view the broader ecosystems, leveraging efforts to build capacity for ecosystem restoration and improving our understanding of watershed health. We recognize that the health of life - such as our own - and the health of our surroundings are interconnected, with our overall well-being dependent on the health of our environment.


<br>

```{r previous-work-intro, results='asis'}
if (params$project_region == "peace") {

  cat("
This project builds on prior years of Society for Ecosystem Restoration Northern BC (SERNbc) fish passage reporting [@irvine2020ParsnipRiver; @irvine2022PEAF22F3577DCARestoring; @irvine_winterscheidt2023RestoringFish; @irvine_winterscheidt2024RestoringFish; @irvine_schick2025RestoringFish]; per-site links to the relevant prior reports are in the [Assessment Data Summary](#app-site-data) appendix.
")

} else if (params$project_region == "fraser") {

  cat("
This project builds on prior years of Society for Ecosystem Restoration Northern BC (SERNbc) fish passage reporting [@irvine_schick2024Restoringfish]; per-site links to the relevant prior reports are in the [Assessment Data Summary](#app-site-data) appendix.
")

} else if (params$project_region == "skeena") {

  cat("
This project builds on prior years of Society for Ecosystem Restoration Northern BC (SERNbc) fish passage reporting [@irvine2021BulkleyRiver; @irvine2022BulkleyRiver; @irvine_etal2023BulkleyWatershed; @irvine_wintersheidt2023SkeenaWatershed; @irvine_schick2023SkeenaWatershed; @irvine_schick2025SkeenaWatershed]; per-site links to the relevant prior reports are in the [Assessment Data Summary](#app-site-data) appendix.
")

}
```

<br>

The health and viability of freshwater fish populations can depend on access to tributary and off channel areas which
provide refuge during high flows, opportunities for foraging, overwintering habitat, spawning habitat and summer rearing
habitat [@bramblett_etal2002SeasonalUse; @swales_levings1989RoleOffChannel; @diebel_etal2015EffectsRoad].  Culverts can
present barriers to fish migration due to low water depth, increased water velocity, turbulence, a vertical drop at the
culvert outlet and/or maintenance issues [@slaney_zaldokas1997FishHabitat; @cote_etal2005Fishpassage]. As road
crossing structures are commonly upgraded or removed there are numerous opportunities to restore connectivity by
ensuring that fish passage considerations are incorporated into repair, replacement, relocation and deactivation
designs.

<br>

Although remediation and replacement of stream crossing structures can have benefits to local fish populations, the costs of remedial works can be significant and the impacts of the work often complex to evaluate and quantify. Additionally, allocation of ecosystem restoration funding towards infrastructure upgrades on transportation right of ways are not always considered ethical under all circumstances from all perspectives. When funds are finite and invested groups are engaged in fund raising, cost benefits and the ethics of crossing replacements should be explored collaboratively alongside the cost benefits and ethics of alternative investment activities including transportation corridor relocation/deactivation, land procurement/covenant, cattle exclusion, riparian/floodplain restoration, habitat complexing, water conservation, commercial/recreational fishing management, salt water interventions and research.

<!-- <br> -->

<!-- Through the ongoing development of open source analysis, data presentation and project collaboration tools we are  -->
<!-- identifying new restoration opportunities, clarifying restoration benefits, communicating with the broader community and  -->
<!-- implementing on the ground works.  -->

<br>

Please note that at the time of reporting this document can be considered a living document. See the [Open Source Reporting Framework](#open-source-reporting-framework) section for details on version tracking and project documentation.

