# Results and Discussion

## Engage Partners

Project advancement in `r params$project_year` depended on engagement with partners across multiple agencies, Nations, industrial tenure holders, funders, and collaborators. Touchpoints during the year (email, video calls, meetings, field-based coordination, and contract administration) are summarised in Table \@ref(tab:tab-engage-partners).

```{r tab-engage-partners}
readr::read_csv("data/communications/partners_2025.csv", show_col_types = FALSE) |>
  dplyr::arrange(affiliation) |>
  dplyr::select(`Organization / Role` = affiliation, `Topic / Focus` = topic) |>
  fpr::fpr_kable(
    caption_text = "Project partners engaged, May 2025 - May 2026.",
    scroll = gitbook_on
  )
```

<br>

### McLeod Lake Indian Band

The McLeod Lake Indian Band (MLIB) is a core collaborator in this work. In `r params$project_year`, MLIB technicians were integrated into all aspects of the program — data transcription, eDNA filtration, backpack electrofishing, fish processing, and the routine effectiveness evaluations at both baseline and remediated crossings — working alongside the field crew throughout Tse'khene territory.

<br>

Capacity building was a deliberate part of this. The University of Northern British Columbia (UNBC) delivered an electrofishing training course hosted in the community, funded in part through this project alongside other funding sources, equipping community members with backpack-electrofishing skills applicable to fisheries work across their territory.

<br>

Field data collection spanned more than ten sites across two systems — tributaries to the Table River and tributaries to Kerry Lake. Work included backpack electrofishing with fish lengths and weights recorded and PIT-tag surgical implantation, alongside extensive environmental DNA (eDNA) collection. This fed both the program-scale eDNA effort ([Appendix - Environmental DNA Results](#app-edna)) and the multi-year effectiveness-monitoring baselines at the Table River and Kerry Lake sites.

<br>

`r if(gitbook_on){knitr::asis_output("Results of Phase 1 and Phase 2 assessments are summarized in Figure \\@ref(fig:map-interactive) with additional details provided in sections below.")}`

<br>

```{r map-interactive, fig.cap='Map of fish passage and habitat confirmation results', eval=gitbook_on}

##make colors for the priorities
pal <- 
   leaflet::colorFactor(palette = c("red", "yellow", "grey", "black"), 
               levels = c("High", "Moderate", "Low", "No Fix"))

pal_phase1 <- 
   leaflet::colorFactor(palette = c("red", "yellow", "grey", "black"), 
               levels = c("High", "Moderate", "Low", NA))

  
map <- leaflet::leaflet(height=500, width=780) |>
  leaflet::addTiles()  |> 
  # leafem::addMouseCoordinates(proj4 = 26911) |> ##can't seem to get it to render utms yet
  # leaflet::addProviderTiles(providers$"Esri.DeLorme") |> 
  leaflet::addProviderTiles("Esri.WorldTopoMap", group = "Topo") |>
  leaflet::addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") |> 
  leaflet::addPolygons(data = wshd_study_areas, color = "#F29A6E", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0,
              fillColor = "#F29A6E", label = wshd_study_areas$watershed_group_name) |>
  leaflet::addPolygons(data = wshds, color = "#0859C6", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0.25,
              fillColor = "#00DBFF",
              label = wshds$stream_crossing_id,
              popup = leafpop::popupTable(x = dplyr::select(wshds |> sf::st_set_geometry(NULL),
                                                     Site = stream_crossing_id,
                                                     elev_site:area_km),
                                          feature.id = F,
                                          row.numbers = F),
              group = "Phase 2") |> 
  leaflet::addLegend(
    position = "topright",
    colors = c("red", "yellow", "grey", "black"),
    labels = c("High", "Moderate", "Low", 'No fix'), opacity = 1,
    title = "Fish Passage Priorities") |> 
  
leaflet::addCircleMarkers(data=dplyr::filter(tab_map_phase_1, assess_type_phase1 == "Yes" | assess_type_reassessment == "Yes"),
                 label = dplyr::filter(tab_map_phase_1, assess_type_phase1 == "Yes" | assess_type_reassessment == "Yes") |>
                   dplyr::pull(pscis_crossing_id),
                 # label = tab_map_phase_1$pscis_crossing_id,
                 labelOptions = leaflet::labelOptions(noHide = F, textOnly = TRUE),
                 popup = leafpop::popupTable(x = dplyr::select((tab_map_phase_1 |> sf::st_set_geometry(NULL) |> dplyr::filter(assess_type_phase1 == "Yes" | assess_type_reassessment == "Yes")),
                                                        Site = pscis_crossing_id, Priority = priority_phase1, Stream = stream_name, Road = road_name, `Habitat value`= habitat_value, `Barrier Result` = barrier_result, `Culvert data` = data_link, `Culvert photos` = photo_link, `Model data` = model_link),
                                             feature.id = F,
                                             row.numbers = F),
                 radius = 9,
                 fillColor = ~pal_phase1(priority_phase1),
                 color= "#ffffff",
                 stroke = TRUE,
                 fillOpacity = 1.0,
                 weight = 2,
                 opacity = 1.0,
                 group = "Phase 1") |> 
  leaflet::addPolylines(data=habitat_confirmation_tracks,
               opacity=0.75, color = '#e216c4',
               fillOpacity = 0.75, weight=5, group = "Phase 2") |>

leaflet::addAwesomeMarkers(
    lng = as.numeric(photo_metadata$gps_longitude),
    lat = as.numeric(photo_metadata$gps_latitude),
    popup = leafpop::popupImage(photo_metadata$url, src = "remote"),
    clusterOptions = leaflet::markerClusterOptions(),
    group = "Phase 2") |>


  leaflet::addCircleMarkers(
    data=tab_map_phase_2,
    label = tab_map_phase_2$pscis_crossing_id,
    labelOptions = leaflet::labelOptions(noHide = T, textOnly = TRUE),
    popup = leafpop::popupTable(x = dplyr::select((tab_map_phase_2 |> sf::st_drop_geometry()),
                                           Site = pscis_crossing_id,
                                           Priority = priority,
                                           Stream = stream_name,
                                           Road = road_name,
                                           `Habitat (m)`= upstream_habitat_length_m,
                                           Comments = comments,
                                           `Culvert data` = data_link,
                                           `Culvert photos` = photo_link,
                                           `Model data` = model_link),
                                feature.id = F,
                                row.numbers = F),
    radius = 9,
    fillColor = ~pal(priority),
    color= "#ffffff",
    stroke = TRUE,
    fillOpacity = 1.0,
    weight = 2,
    opacity = 1.0,
    group = "Phase 2"
    ) |>

  leaflet::addLayersControl(
    baseGroups = c(
      "Esri.DeLorme",
      "ESRI Aerial"),
    overlayGroups = c("Phase 1", "Phase 2"),
    options = leaflet::layersControlOptions(collapsed = F)) |>
  leaflet.extras::addFullscreenControl() |>
  leaflet::addMiniMap(tiles = leaflet::providers$"Esri.NatGeoWorldMap",
             zoomLevelOffset = -6, width = 100, height = 100)

map



```


## Site Assessment Data

A summary of fish passage assessment procedures conducted through SERNbc since `r params$project_start_year` — including assessments, habitat confirmations, design, remediation, monitoring, fish sampling, eDNA sampling, and drone imagery — is provided in [Appendix - Assessment Data Summary](#app-site-data).


```{r cd-rollup-body, include = FALSE}
cd_data_body <- readRDS("data/gis/cd_peace.rds")
cmp_body     <- cd_data_body$regional$cmp
cmp_pct_body <- cd_data_body$regional$cmp_pct
trn_body     <- cd_data_body$regional$trn
pick <- function(tbl, var, period = "annual") {
  tbl[tbl$variable == var & tbl$period == period, ]
}
tmean_d    <- pick(cmp_body,     "tmean")$difference
tmax_d     <- pick(cmp_body,     "tmax")$difference
tmin_d     <- pick(cmp_body,     "tmin")$difference
vpd_d      <- pick(cmp_body,     "vpd")$difference
prcp_p     <- pick(cmp_pct_body, "prcp")$difference
swe_p      <- pick(cmp_pct_body, "swe")$difference
snowfall_p <- pick(cmp_pct_body, "snowfall")$difference
summer_swe_p   <- pick(cmp_pct_body, "swe",      "summer")$difference
spring_melt_p  <- pick(cmp_pct_body, "snowmelt", "spring")$difference
doy50_d    <- pick(cmp_body, "snowmelt_doy_50")$difference
```

## Climate Departure

Four findings from the regional climate-departure analysis carry to fish-passage prioritisation.

**Warming is broad, fast, and significant region-wide.** All five ecoregions of the FWCP Peace Region warmed about `r sprintf("%.1f", tmean_d)` °C since 1951 — annual mean (+`r sprintf("%.1f", tmean_d)` °C), daytime maximum (+`r sprintf("%.1f", tmax_d)` °C), and overnight minimum (+`r sprintf("%.1f", tmin_d)` °C) all tell the same story. The trend is highly significant (Mann-Kendall p < 0.001 in every ecoregion). A subtle west-of-Rockies gradient layers on top, with the western and central ecoregions running about 0.2 °C warmer than the southeast — see [Appendix - Climate Departure](#app-climate-departure) Figure \@ref(fig:cd-map-tmean).

**The atmosphere is drying despite locally rising precipitation.** Vapour pressure deficit rose `r sprintf("%.2f", vpd_d)` hPa region-wide, significant in every ecoregion. Annual precipitation rose `r sprintf("%+.0f", prcp_p)` % regionally but the long-term trend is significant only in the two northernmost ecoregions; soil moisture is essentially flat everywhere because the warmer atmosphere is drinking the extra water back.

**The snowpack signal is about timing, not quantity.** Total annual snowfall is essentially unchanged (`r sprintf("%+.0f", snowfall_p)` %); annual snow water equivalent is down `r sprintf("%.0f", -swe_p)` %. But the seasonal redistribution is sharp — summer snow water equivalent collapsed `r sprintf("%.0f", -summer_swe_p)` %, spring snowmelt rose `r sprintf("%+.0f", spring_melt_p)` %, and the snowmelt midpoint shifted `r sprintf("%.0f", -doy50_d)` days earlier. The freshet is arriving weeks earlier and the high-elevation snowpack that historically lingered into summer no longer does.

**Day-night asymmetry is present.** Overnight minimums warmed `r sprintf("%.1f", tmin_d)` °C while daytime maximums warmed `r sprintf("%.1f", tmax_d)` °C — overnight outpacing daytime is the textbook fingerprint of greenhouse warming [@karl_etal1993NewPerspective]. Mean diurnal range has narrowed by `r sprintf("%.1f", tmin_d - tmax_d)` °C across the record.

For barrier prioritisation these signals cut both ways. Where upstream reaches are cold-limited, added growing-degree-days can raise the intrinsic potential of habitat above the barrier and *increase* the value of restoring passage; where upstream reaches sit closer to the upper edge of the cold-water thermal niche, the same warming reduces usable habitat and lowers the payoff. The watershed-group breakdown in [Appendix - Climate Departure](#app-climate-departure) maps the five-ecoregion signal onto the 16 FWCP Peace watershed groups.

## Planning — Habitat, Connectivity and Floodplain Modelling

### Habitat Modelling
Habitat modelling from `bcfishpass` including access model, linear spawning/rearing habitat model and lateral habitat
connectivity models for watershed groups within our study area were updated for the spring of 2025 and are included
spatially in the collaborative GIS project.  A snapshot of these outputs related to each modeled and PSCIS stream
crossing structure are also included within an `sqlite` database within this year's project reporting/code repository [here](https://github.com/NewGraphEnvironment/fish_passage_peace_2024_reporting/blob/main/data/bcfishpass.sqlite).


#### Statistical Support for `bcfishpass` Fish Habitat Modelling Updates

The statistical-modelling methodology — channel-width prediction, stream-temperature SSN modelling with Poisson Consulting [@hill_etal2024Spatialstream; @hill_etal2025Spatialstream], the broader `fresh` + `link` habitat-modelling framework, and the calibration trajectory against `bcfishobs` and research activities targeting known high-value spawning and rearing — is described in the Methods chapter under Statistical Support for Habitat Modelling. The outputs feeding intrinsic-habitat and connectivity prioritization are summarised below under Statistical Habitat Modelling Outputs.



```{r flood-rollup-body, include = FALSE}
gpkg_fp    <- "data/gis/pars.gpkg"
aoi_fp     <- sf::st_read(gpkg_fp, layer = "aoi", quiet = TRUE)
streams_fp <- sf::st_read(gpkg_fp, layer = "streams", quiet = TRUE)
wb_fp      <- sf::st_read(gpkg_fp, layer = "waterbodies", quiet = TRUE)
fp_poly    <- sf::st_read(gpkg_fp, layer = "floodplain", quiet = TRUE)

aoi_fp     <- sf::st_make_valid(aoi_fp)
streams_fp <- suppressWarnings(sf::st_intersection(streams_fp, aoi_fp))
fp_poly    <- suppressWarnings(sf::st_intersection(fp_poly, aoi_fp))

aoi_area_km2    <- as.numeric(sum(sf::st_area(aoi_fp))) / 1e6
fp_area_ha      <- as.numeric(sum(sf::st_area(fp_poly))) / 1e4
fp_pct_aoi      <- 100 * fp_area_ha / (aoi_area_km2 * 100)

streams_in_fp    <- suppressWarnings(sf::st_intersection(streams_fp, fp_poly))
streams_km_total <- as.numeric(sum(sf::st_length(streams_fp))) / 1e3
streams_km_in_fp <- as.numeric(sum(sf::st_length(streams_in_fp))) / 1e3

wb_in_fp    <- suppressWarnings(sf::st_intersection(wb_fp, fp_poly))
lakes_n     <- sum(wb_in_fp$waterbody_type == "L", na.rm = TRUE)
wetlands_n  <- sum(wb_in_fp$waterbody_type == "W", na.rm = TRUE)
```

### Floodplain Delineation

Modelled functional floodplain in the Parsnip River Watershed Group covers `r format(round(fp_area_ha), big.mark = ",")` ha — `r sprintf("%.1f", fp_pct_aoi)` % of the `r format(round(aoi_area_km2), big.mark = ",")` km² watershed group — attached to `r format(round(streams_km_in_fp), big.mark = ",")` km of bull trout accessible order-3+ streams. Within that footprint sit `r lakes_n` lakes and `r format(wetlands_n, big.mark = ",")` mapped wetland polygons — the off-channel inventory that crossing remediations in these reaches stand to reconnect or leave severed, depending on whether design accounts for the floodplain context around the structure. The full metric table, watershed-scale map, and south-east detail map are in [Appendix - Floodplain Delineation](#app-floodplain).


## Statistical Habitat Modelling Outputs

Connectivity modelling now runs weekly with automated updates across all `r length(params$wsg_code)` watershed groups of the FWCP Peace area (the Upper Peace River group falls only partially within the FWCP boundary), driven by `bcfishpass` outputs using channel width, gradient, and provincial / federal barrier inventories — channel-width estimates are themselves a product of the Bayesian statistical work described in the methods [@thorley_irvine2021ChannelWidtha; @thorley_etal2021ChannelWidthc]. Field assessments were completed this year in five of those watershed groups (Parsnip River, Crooked River, Carp Lake, Nation River, and Parsnip Arm) to maintain momentum and surface candidate sites and partners. The work has the support of the Provincial Fish Passage Technical Working Group (FPTWG) — a partnership between the Province of British Columbia (WLRS, MOF, MOE, MOTI) and Fisheries and Oceans Canada.

<br>

Alongside the weekly automated updates, we use our open-source R packages `fresh` and `link` to expand, experiment with, and develop new project-specific habitat-modelling outputs. The first concrete application is an Arctic grayling intrinsic-habitat and connectivity model for the FWCP Peace area, parameterized on stream size and gradient, with further covariates such as stream temperature ([`water-temp-bc`](https://www.newgraphenvironment.com/water-temp-bc/)) available to incorporate as the framework develops. We present our first outputs — reproducing `bcfishpass`'s bull-trout classification and extending it to Arctic grayling for the Parsnip River Watershed Group — in [Appendix - Parsnip River Habitat and Connectivity Modelling](#app-habitat-connectivity) (Figures \@ref(fig:link-map-bt)–\@ref(fig:link-map-detail)).

<br>


## Fish Passage Assessments

Field assessments were conducted between `r format(min(form_pscis$date), format="%B %d, %Y")` and `r format(max(form_pscis$date), format="%B %d, %Y")` by Allan Irvine, R.P.Bio. and Lucy Schick, B.Sc. A total of `r form_pscis |> filter(assess_type_phase1 == "Yes"|assess_type_reassessment == "Yes") |> nrow()` Fish Passage Assessments were completed, including `r n_distinct(pscis_phase1$my_crossing_reference)` Phase 1 assessments and `r pscis_reassessments |> nrow()` reassessments.

<br>

Of the `r form_pscis |> filter(assess_type_phase1 == "Yes"|assess_type_reassessment == "Yes") |> nrow()` sites where fish passage assessments were completed, `r pscis_phase1 |> nrow()` were not yet inventoried in the PSCIS system.  This included `r pscis_phase1 |> filter(barrier_result == 'Passable') |> nrow()` crossings considered "passable", `r pscis_phase1 |> filter(barrier_result == 'Potential') |> nrow() |> english::as.english()` crossings considered a "potential" barrier, and `r pscis_phase1 |> filter(barrier_result == 'Barrier') |> nrow()` crossings were considered "barriers" according to threshold values based on culvert embedment, outlet drop, slope, diameter (relative to channel size) and length [@moe2011Fieldassessment].

<!-- Additionally, although all were considered fully passable, `r pscis_phase1 |> filter(barrier_result == 'Unknown') |> nrow() |> english::as.english()` crossings assessed were fords and ranked as "unknown" according to the provincial protocol. -->

<!-- <br> -->

<!-- Reassessments were completed at `r pscis_reassessments |> nrow()` sites where PSICS data required updating. -->

<br>

A summary of crossings assessed, a rough cost estimate for remediation, and a priority ranking for follow-up for Phase 1 sites is presented in Table \@ref(tab:cost-est-phase-1). Detailed data with photos are presented in `r if(gitbook_on){knitr::asis_output("[Appendix - Phase 1 Fish Passage Assessment Data and Photos]")}else(knitr::asis_output("[Attachment - Phase 1 Data and Photos](#attach-pdf-phase1-dat)"))`.


<br>



`r if(gitbook_on){knitr::asis_output("")} else knitr::asis_output("<br><br><br><br>")`

<br>


```{r cost-est-phase-1, eval = T}

 tab_cost_est_phase1 |>
  select(`PSCIS ID`:`Cost Est ( $K)`) |> 
   fpr::fpr_kable(caption_text = paste0("Upstream habitat estimates and cost benefit analysis for Phase 1 assessments ranked as a 'barrier' or 'potential' barrier. ", sp_network_caption),
                  scroll = T)
 
```



## Habitat Confirmation Assessments

During the `r params$project_year` field assessments, habitat confirmation assessments were conducted at `r form_fiss_site |> dplyr::filter(is.na(ef)) |> dplyr::distinct(site) |> nrow()` sites within the `r wsg_names |> knitr::combine_words()`  watershed groups. A total of approximately `r habitat_confirmations_priorities |> pull(length_surveyed) |> sum(na.rm = TRUE) |> round(-3)/1000`km of stream was assessed. 


<br>

As collaborative decision-making was ongoing at the time of reporting, site prioritization can be considered preliminary. Results are summarized in `r if(gitbook_on){knitr::asis_output("Figure \\@ref(fig:map-interactive) and ")}` Table \@ref(tab:tab-overview) with raw habitat `r if (exists("fish_data_complete")) "and fish sampling " else ""` data included in [Attachment - Data](#attach-dat). A summary of preliminary modelling results illustrating quantities of `r stringr::str_to_lower(model_species_name)` spawning and rearing habitat potentially available upstream of each crossing as estimated by measured/modelled channel width and upstream accessible stream length are presented in Figure \@ref(fig:plot-model-all). Detailed information for each site assessed with Phase 2 assessments (including maps) is presented in the per-site habitat-confirmation appendices: [Tributary to Parsnip River (PSCIS 199663)](#tributary-to-parsnip-river---199663---appendix), [Tributary to Nation River (PSCIS 203597)](#tributary-to-nation-river---203597---appendix), and [Tributary to Williston Reservoir (PSCIS 203605)](#tributary-to-williston-reservoir---203605---appendix).

`r if(gitbook_on){knitr::asis_output("")} else knitr::asis_output("<br><br><br><br>")`

<br>


```{r tab-overview, eval = TRUE}
table_phase2_overview <- function(dat, caption_text = '', font = font_set, scroll = TRUE){
  dat2 <- dat |>
    kable(caption = caption_text, booktabs = T, label = NA) |>
    kableExtra::kable_styling(c("condensed"),
                              full_width = T,
                              font_size = font) |>
    kableExtra::column_spec(column = c(11), width_min = '1.5in') |>
    kableExtra::column_spec(column = c(1:10), width_max = '1in')

  if(identical(scroll,TRUE)){
    dat2 <- dat2 |>
      kableExtra::scroll_box(width = "100%", height = "500px")
  }
  dat2
}

tab_overview |> 
  table_phase2_overview(caption_text = paste0("Overview of habitat confirmation sites. ", sp_rearing_caption), 
                        scroll = gitbook_on)

```

`r if(gitbook_on){knitr::asis_output("<br>")} else knitr::asis_output("\\pagebreak")`


```{r phase-2-cv-assess-sum}
fpr::fpr_table_cv_summary(dat = pscis_phase2)  |>  
  fpr::fpr_kable(caption_text = 'Summary of Phase 2 fish passage reassessments.', scroll = F)
```


`r if(gitbook_on){knitr::asis_output("<br>")} else knitr::asis_output("\\pagebreak")`


```{r cost-est-phase-2, eval=T}
tab_cost_est_phase2 |>
    dplyr::rename(
    `PSCIS ID` = pscis_crossing_id,
    Stream = stream_name,
    Road = road_name,
    `Barrier Result` = barrier_result,
    `Habitat value` = habitat_value,
    `Habitat Upstream (m)` = upstream_habitat_length_m,
    `Stream Width (m)` = avg_channel_width_m,
    Fix = crossing_fix_code,
    `Cost Est ( $K)` = cost_est_1000s,
    `Cost Benefit (m / $K)` = cost_net,
    `Cost Benefit (m2 / $K)` = cost_area_net
  ) |> 
  fpr::fpr_kable(caption_text = paste0("Cost benefit analysis for Phase 2 assessments. ", sp_rearing_caption),
                 scroll = F)

```


`r if(gitbook_on){knitr::asis_output("<br>")} else knitr::asis_output("\\pagebreak")`

```{r tab-habitat-summary}
tab_hab_summary |> 
  # `Location` is capitalised upstream ("Upstream"/"Downstream"); match it exactly.
  # stringr 1.6.0 made `str_like()` case sensitive, so a lowercase 'upstream'
  # pattern silently matched nothing and emptied this table.
  dplyr::filter(Location == "Upstream") |>
  dplyr::select(-Location) |> 
  dplyr::rename(`PSCIS ID` = Site, `Length surveyed upstream (m)` = `Length Surveyed (m)`) |>
  fpr::fpr_kable(caption_text = 'Summary of Phase 2 habitat confirmation details.', scroll = F)


```

<br>

```{r tab-wshd-sum-all}
fpr::fpr_table_wshd_sum() |> 
  fpr::fpr_kable(caption_text = paste0('Summary of watershed area statistics upstream of Phase 2 crossings.'),
                                 footnote_text = 'Elev P60 = Elevation at which 60% of the watershed area is above', scroll = F)

```

<br>

```{r plot-model-all-cap}

my_caption = paste0("Summary of potential rearing and spawning habitat upstream of habitat confirmation assessment sites. ", model_species_name," rearing and spawning models used for habitat estimates (total length of stream network <", rear_gradient, "% and <", spawn_gradient, "% gradient, respectively).")

```


```{r plot-model-all, fig.cap= my_caption, eval=T}


bcfp_xref_plot <- xref_bcfishpass_names |> 
  filter(!is.na(id_join) &
         !stringr::str_detect(bcfishpass, 'below') &
         !stringr::str_detect(bcfishpass, 'all') &
         !stringr::str_detect(bcfishpass, '_ha') &
         (stringr::str_detect(bcfishpass, 'rearing') | 
          stringr::str_detect(bcfishpass, 'spawning')))


bcfishpass_phase2_plot_prep <- bcfishpass |> 
  dplyr::mutate(dplyr::across(where(is.numeric), round, 1)) |>
  dplyr::filter(stream_crossing_id %in% (pscis_phase2 |> dplyr::pull(pscis_crossing_id))) |> 
  dplyr::select(stream_crossing_id, dplyr::all_of(bcfp_xref_plot$bcfishpass)) |> 
  dplyr::mutate(stream_crossing_id = as.factor(stream_crossing_id)) |> 
  tidyr::pivot_longer(cols = bt_rearing_km:bt_spawning_km) |> 
  dplyr::filter(value > 0.0 &
                !is.na(value)) |> 
  dplyr::mutate(
    name = dplyr::case_when(stringr::str_detect(name, '_rearing') ~ paste0(model_species_name, " rearing km"),
                            TRUE ~ name),
    name = dplyr::case_when(stringr::str_detect(name, '_spawning') ~ paste0(model_species_name, " spawning km"),
                            TRUE ~ name)
    # Use when more than one modelling species 
   # name = stringr::str_replace_all(name, '_rearing', ' rearing'),
   # name = stringr::str_replace_all(name, '_spawning', ' spawning')
  )
  
  
  
bcfishpass_phase2_plot_prep |> 
  ggplot2::ggplot(ggplot2::aes(x = stream_crossing_id, y = value)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~name, ncol = 2) +
  # ggdark::dark_theme_bw(base_size = 11) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1)) +
  ggplot2::labs(x = "Site", y = "Modelled habitat (km)")


```


### Fish Sampling

Fish sampling was conducted at `r tab_fish_sites_sum |> n_distinct()` sites within 6 streams, with a total of `r tab_fish_summary |> dplyr::pull(count_fish) |> sum(na.rm = TRUE)` fish captured. Fork length, weight, and species were documented for each fish. Salmonids with fork lengths >60mm were PIT-tagged to facilitate long-term tracking of health and movement. Fork length data was used to delineate salmonids based on life stages: fry (0 to 65mm), parr (>65 to 110mm), juvenile (>110mm to 140mm), and adult (>140mm) by visually assessing the histograms presented in Figure \@ref(fig:fish-histogram). A summary of sites assessed is included in Table \@ref(tab:tab-fish-site-summary), and raw data is provided in [Attachment - Data](#attach-dat). A summary of density results is also presented in Figure \@ref(fig:plot-fish-all).


```{r fish-sampling-text, eval = F}


# 
# <br>
# 
# Results are presented in detail within the individual appendices in this report with the [2023 report](https://newgraphenvironment.github.io/fish_passage_peace_2023_reporting) updated with fish sampling data for PSCIS crossing 125131 on Chuchinka-Table FSR.  Documentation for each site can be accessed `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("online within the Results and Discussion section of the report found ", ngr::ngr_str_link_url(url_base = params$report_url, anchor_text = "here")))}else{knitr::asis_output("by searching the site number in Table \\@ref(tab:tab-sites-cap) and clicking the _Link Report_")}`.
# 
# <br>
# 
# `r if(gitbook_on){knitr::asis_output("")} else knitr::asis_output("<br><br><br><br>")`
# 
# <br>


```


```{r tab-fish-site-summary, eval=TRUE}
tab_fish_sites_sum |>
  fpr::fpr_kable(caption_text = 'Summary of electrofishing sites.', scroll = FALSE)
```

<br>

```{r plot-fish-all, fig.cap='Boxplots of densities (fish/100m2) of fish captured downstream (ds) and upstream (us) by electrofishing.', eval=TRUE}

plot_fish_box_all <- fish_abund |> 
  dplyr::filter(
    !species_code %in% c('Mountain Whitefish', 
                         'Sucker (General)',
                         'NFC',
                         'Longnose Sucker',
                         'Sculpin (General)')
  ) |> 
  ggplot2::ggplot(ggplot2::aes(x = location, y = density_100m2)) +
  ggplot2::geom_boxplot(fill = NA) +
  ggplot2::facet_grid(life_stage ~ site, scales = "free_y", as.table = TRUE) +

  # ggplot2::facet_grid(site ~ species_code, scales = "fixed", as.table = TRUE) +
  ggplot2::theme(legend.position = "none", axis.title.x = ggplot2::element_blank()) +
  ggplot2::geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 1) +
  ggplot2::ylab(expression(Density ~ (Fish/100 ~  m^2))) +
  ggplot2::theme_bw()
  # ggdark::dark_theme_bw()



plot_fish_box_all
```

<br>

```{r fish-histogram, fig.cap= 'Histograms of fish lengths by species.  Fish captured by electrofishing during habitat confirmation assessments.', eval = TRUE}

# Fish histogram -----------------------------------------------------------------------
bin_1 <- floor(min(fish_data_complete$length, na.rm = TRUE) / 5) * 5
bin_n <- ceiling(max(fish_data_complete$length, na.rm = TRUE) / 5) * 5
bins <- seq(bin_1, bin_n, by = 5)

# Check what species we have and filter out any we don't want
# fish_data_complete |> dplyr::distinct(species)

plot_fish_hist <- ggplot2::ggplot(
  fish_data_complete |> dplyr::filter(!species %in% c('Fish Unidentified Species', 'Sculpin (General)', 'NFC')),
  ggplot2::aes(x = length)
) +
  ggplot2::geom_histogram(breaks = bins, alpha = 0.75,
                          position = "identity", size = 0.75) +
  ggplot2::labs(x = "Fork Length (mm)", y = "Count (#)") +
  ggplot2::facet_wrap(~species) +
  ggplot2::theme_bw(base_size = 8) +
  ggplot2::scale_x_continuous(breaks = bins[seq(1, length(bins), by = 2)]) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

plot_fish_hist


```


### Environmental DNA (eDNA) sampling

```{r tab-edna-summary-prep, include=FALSE}
# Phase 2 (issue #6): load Peace-filtered eDNA inputs (snapshotted from
# template repo via scripts/edna_inputs_snapshot.R) and compute the
# per-species summary used in the table below.

norm_lgl <- function(x) {
  if (is.logical(x)) return(x)
  out <- toupper(as.character(x)) == "TRUE"
  out[is.na(x)] <- FALSE
  out
}

# Used in main Results, the thematic appendix, and the per-site appendices —
# defining once here so every consumer downstream in the bookdown merge order
# can use it without redefinition.
fmt_targets <- function(target, max_pos, retested) {
  if (length(target) == 0) return("—")
  parts <- paste0(
    target,
    "(", ifelse(is.na(max_pos), "?", max_pos), ")",
    ifelse(retested, "*", "")
  )
  paste(parts, collapse = ", ")
}

edna_bystargets_peace <- readr::read_csv(
  "data/edna_unbc_results_2025_by_site_target.csv",
  show_col_types = FALSE
) |>
  dplyr::filter(grepl("sern_peace_fwcp_2023", source)) |>
  dplyr::mutate(
    control_blank_field           = norm_lgl(control_blank_field),
    control_blank_office          = norm_lgl(control_blank_office),
    control_species_present_field = norm_lgl(control_species_present_field)
  )

# Real environmental samples = exclude both office blanks (fake coords) and
# field blanks (QA samples, not site eDNA — covered separately on the map's
# Controls layer).
edna_real_peace <- edna_bystargets_peace |>
  dplyr::filter(!control_blank_office, !control_blank_field)

n_edna_real_sites    <- dplyr::n_distinct(edna_real_peace$site_id)
n_edna_field_blanks  <- edna_bystargets_peace |>
  dplyr::filter(control_blank_field)  |> dplyr::distinct(site_id) |> nrow()
n_edna_office_blanks <- edna_bystargets_peace |>
  dplyr::filter(control_blank_office) |> dplyr::distinct(site_id) |> nrow()

# UNBC assay code -> common name
edna_species_names <- c(
  BULT  = "Bull Trout",
  GRAY  = "Arctic Grayling",
  RAIN  = "Rainbow Trout",
  RAIN2 = "Rainbow Trout (confirmation primer)",
  CHIN  = "Chinook Salmon",
  COHO  = "Coho Salmon",
  # In Peace tributaries above the Peace Canyon Dam there are no
  # anadromous Sockeye Salmon — the SOCK ddPCR assay detects landlocked
  # Kokanee. Different from Fraser/Skeena where SOCK = Sockeye Salmon.
  SOCK  = "Kokanee",
  BURB  = "Burbot"
)

edna_summary_peace <- edna_real_peace |>
  dplyr::group_by(target) |>
  dplyr::summarise(
    sites_tested        = dplyr::n_distinct(site_id),
    sites_detected      = dplyr::n_distinct(site_id[any_clean]),
    sites_subthreshold  = dplyr::n_distinct(site_id[any_crude & !any_clean]),
    sites_no_detection  = dplyr::n_distinct(site_id[!any_crude]),
    .groups = "drop"
  ) |>
  dplyr::mutate(Species = dplyr::coalesce(edna_species_names[target], target)) |>
  dplyr::select(
    Species,
    `Code`                              = target,
    `Sites tested`                      = sites_tested,
    `Detected (≥4 droplets)`           = sites_detected,
    `Sub-threshold only (1-3 droplets)` = sites_subthreshold,
    `No detection (0 droplets)`         = sites_no_detection
  ) |>
  dplyr::arrange(dplyr::desc(`Detected (≥4 droplets)`), Species)
```

New in 2025, environmental DNA (eDNA) sampling was incorporated into the program at both previously assessed and newly added sites. A total of `r form_edna |> nrow()` collection events were undertaken across `r form_edna |> dplyr::distinct(stream_name) |> nrow()` streams, including `r n_edna_real_sites` real environmental samples, `r n_edna_field_blanks` field blanks, and `r n_edna_office_blanks` office blanks for quality assurance.

Per-species detection results across the `r n_edna_real_sites` real environmental samples are summarized in Table \@ref(tab:tab-edna-summary). Detections are based on the standard ddPCR call threshold of ≥4 positive droplets; sub-threshold signals (1-3 droplets) are tracked separately as they are not interpretable as species presence per ddPCR convention. Site-level detail (per-site detection table, field blanks, and retests) is provided in [Appendix - Environmental DNA Results](#app-edna), and the `r ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "edna_unbc_results_2025_peace_map.html", anchor_text = "interactive Peace eDNA map")` shows toggleable per-species layers.

<br>

```{r tab-edna-summary}
edna_summary_peace |>
  fpr::fpr_kable(
    caption_text = "Per-species summary of eDNA results across real environmental samples in the Peace project area (excludes field and office blanks).",
    scroll = FALSE
  )
```

<br>

## Engineering Design

In `r params$project_year`, the engineering-design pipeline shifted in two ways. The proposed Kerry Lake FSR structure replacement at PSCIS 198692, originally scoped for 2024/25, was shelved after Sinclar Forest Products and project partners agreed that road decommissioning and structure removal at both PSCIS 198692 (km 9) and PSCIS 198693 (km 8) is a better outcome on both economic and ecological grounds; Sinclar will cover removal and deactivation to minimum provincial standards in 2025/26 (baseline-monitoring context in [Appendix - Tributary to Kerry Lake](#trib-to-kerry-2025)).

BC Timber Sales (BCTS) shared road-deactivation plans for the Tributary to Nation River crossings deactivated in fall 2025 — used to scope above-minimum-standards riparian restoration (soil decompaction, riparian planting, coarse woody debris placement) for 2026/27. BCTS is also lining up Kennedy Siding (PSCIS 199663) on the Chuchinka-Colbourne FSR for a site plan and engineering design in spring 2026, with the project providing modelled habitat and connectivity context to inform the design scope.

All sites with a design for remediation of fish passage can be found `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("online within the Results and Discussion section of the report found ", ngr::ngr_str_link_url(url_base = params$report_url, anchor_text = "here")))}else{knitr::asis_output("in Table \\@ref(tab:tab-sites-cap) by filtering using Design = yes")}`.


```{r remediations-text, eval=F}

# ## Remediations
# 
# In 2024, remediation of fish passage was completed at PSCIS crossing 125231, located on Tributary to Table River at km 21 
# on the Chuchinka-Table FSR. The crossing was replaced with a clear-span bridge by Canfor with environmental oversight and engineering from DWB Consulting Services Ltd. Half the total funding for the project was provided by FWCP through coordination from SERNbc. More information regarding the crossing can be found within this report [here](https://www.newgraphenvironment.com/fish_passage_peace_2023_reporting/tributary-to-the-table-river---125231---appendix.html).  
# 
# <br>
# 
# All sites where remediation of fish passage has been completed can be found `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("online within the Results and Discussion section of the report found ", ngr::ngr_str_link_url(url_base = params$report_url, anchor_text = "here")))}else{knitr::asis_output("in Table \\@ref(tab:tab-sites-cap) by filtering using Remediation = yes")}`. 

```


## Monitoring

In `r params$project_year`, effectiveness monitoring was conducted at two previously remediated crossings, and baseline data collection (fish sampling and eDNA only — no effectiveness form) was completed at one site ahead of planned remediation:

- **Tributary to Missinka River (PSCIS 125179)** on Chuchinka-Missinka FSR — effectiveness monitoring (form + fish sampling + eDNA), extending the multi-year baseline established in @irvine_winterscheidt2023RestoringFish and @irvine_schick2025RestoringFish. Detail in [Appendix - Tributary to Missinka River](#trib-to-missinka-2025).

- **Tributary to the Table River (PSCIS 125231)** on Chuchinka-Table FSR — effectiveness monitoring (form + fish sampling + eDNA) following the 2023 replacement of the crossing with a clear-span bridge by Canfor with FWCP funding via SERNbc coordination [@irvine_winterscheidt2024RestoringFish]. Detail in [Appendix - Tributary to the Table River](#trib-to-table-2025).

- **Tributary to Kerry Lake (PSCIS 198692)** — pre-remediation baseline (fish sampling and eDNA) to inform future effectiveness monitoring. The remediation plan has pivoted: where the 2024/25 work envisaged a structure replacement, Sinclar Forest Products — as the tenure holder for the Kerry Lake FSR — has now committed to road decommissioning and removal of both PSCIS 198692 (km 9) and PSCIS 198693 (km 8) in 2025/26 to minimum provincial deactivation standards. The 2024 habitat confirmation is documented in @irvine_schick2025RestoringFish; the 2024–2025 baseline at km 8 (this site) becomes the pre-removal reference for post-removal effectiveness monitoring. Restoration above the minimum deactivation standards (soil decompaction, riparian planting, coarse woody debris placement) is being scoped through the 2026/27 proposal process. Detail in [Appendix - Tributary to Kerry Lake](#trib-to-kerry-2025).

A summary of effectiveness monitoring form metrics across the two effectiveness sites is presented in Table \@ref(tab:tab-mon-overview).

<br>

```{r tab-mon-overview, eval = T}
# started but not sure if this is going to be helpful. 
form_monitoring |> 
 sf::st_drop_geometry() |> 
 dplyr::select(
    `PSCIS ID` = pscis_crossing_id,
    Stream = stream_name,
    Road = road_name,
    `Crossing type` = crossing_type,
    `Crossing subtype` = crossing_subtype,
    `Diameter or span (m)` = diameter_or_span_meters,
    `Length or width (m)` = length_or_width_meters,
    `Assessment comments` = assessment_comment
  ) |> 
   fpr::fpr_kable(caption_text = 'Summary of monitoring sites.', scroll = F)
  

```




## Collaborative GIS Environment

In addition to numerous layers documenting fieldwork activities since `r cat(params$project_start_year)`, a summary of background information spatial layers and tables loaded to the collaborative GIS project (`r params$gis_project_name`) at the time of writing (`r format(lubridate::now(), "%Y-%m-%d")`) is provided in [Appendix - Collaborative GIS Layers](#app-gis).
