---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - Parsnip River Habitat and Connectivity Modelling {-#app-habitat-connectivity}

```{r link-setup, include = FALSE}
library(sf)
```

For any stream, we want to know three things: can a fish get there, is it good
habitat, and what — if anything — blocks the way. We model this for every segment
of the stream network in a watershed group, and for each species we work out:

- **Access** — can the fish reach the segment, or is the route downstream
  likely too steep (gradient), or cut off by a barrier such as a dam,
  waterfall, or perched culvert?
- **Habitat** — if it is reachable, is the segment likely large enough
  (channel width) and its gradient gentle enough for the modelled species to
  use it as **spawning** habitat, **rearing** habitat, or simply passable water
  with neither?
- **Barriers** — the most significant obstacle downstream: a **dam**, a
  **known** barrier that has been assessed in the field (a road culvert, weir,
  etc.), a **modelled** crossing (predicted from road–stream intersections but
  not yet field-checked), or one that has since been **remediated** (fixed).

[bcfishpass](https://github.com/smnorris/bcfishpass) (Hillcrest Geographics)
produces this per-segment classification province-wide. We reproduce it, and
because we re-express the methodology in our own tools we can re-parameterise the
rules to our own values and extend the method to species `bcfishpass` does not
yet model.

We demonstrate both here for the Parsnip River Watershed Group: reproducing the
bull-trout (`BT`) classification, and a net-new classification for Arctic
grayling (`GR`).

The Parsnip River Watershed Group sits between Prince George and Mackenzie, BC.
The Parsnip flows north into the southern arm of Williston Reservoir, joining
the Peace River system; from there the drainage runs Peace → Slave → Mackenzie,
ultimately discharging to the Arctic Ocean via the Mackenzie Delta. Both bull
trout and Arctic grayling are cold-water species whose distributions we resolve
through gradient, channel width, and access. Bull trout — provincially blue-listed and a COSEWIC species of
special concern in the Western Arctic population — spawn as adfluvial migrants
in cold, low-gradient tributaries such as the Misinchinka and Anzac. Arctic
grayling, at the southern edge of their range in the Williston watershed, hold
to cooler, larger (fourth-order and up) clear-water reaches and spawn over fine
gravels.

We build on the canonical [`fwapg`](https://github.com/smnorris/fwapg) /
`bcfishpass` / [`bcfishobs`](https://github.com/smnorris/bcfishobs) stack from
Hillcrest Geographics — porting the processing methodology into our own
[`fresh`](https://github.com/NewGraphEnvironment/fresh) and
[`link`](https://github.com/NewGraphEnvironment/link) packages while still
drawing on the canonical `bcfishpass` inputs (crossing overrides and the like).
That lets us experiment with the rules for species already modelled and extend to
species not yet modelled — here, Arctic grayling — while staying checkable,
segment for segment, against the canonical outputs.

## Modelling parameters {-}

The main parameters are the maximum gradients for access, spawning, and rearing
and a minimum channel width for habitat (we apply further rules too — waterbody
and stream-edge types, clustering of rearing with spawning habitat, and others).
The values we applied are below.

```{r link-params}
param_tab <- data.frame(
  species = c("BT", "GR"),
  access_grad_max = c(0.25, 0.15),
  spawn_grad_max = c(0.0549, 0.0249),
  rear_grad_max = c(0.1049, 0.0349),
  spawn_cw_min_m = c(2, 4),
  rear_cw_min_m = c(1.5, 1.5),
  stringsAsFactors = FALSE
)
knitr::kable(
  param_tab, row.names = FALSE,
  col.names = c("species", "access grad max", "spawn grad max", "rear grad max",
                "spawn CW min (m)", "rear CW min (m)"),
  caption = paste0(
    "Access, spawning, and rearing gradient ceilings and minimum channel ",
    "widths applied for the two species. Grayling's tighter access ceiling and ",
    "gradient windows give it the smaller modelled network."
  )
)
```

```{r link-load, include = FALSE}
gpkg   <- "data/gis/habitat-connectivity.gpkg"
parity <- readRDS("data/gis/habitat-connectivity_parity.rds")

present <- sf::st_layers(gpkg)$name
lyr <- function(name) {
  if (name %in% present) sf::st_read(gpkg, layer = name, quiet = TRUE) else NULL
}

aoi           <- lyr("aoi")
streams       <- lyr("streams")
waterbodies   <- lyr("waterbodies")
named_streams <- lyr("named_streams")
reserves      <- lyr("reserves")
parks         <- lyr("parks")
roads         <- lyr("roads")
railways      <- lyr("railways")

# Context layers (reserves/parks/roads/railways) are pulled by intersection
# with the WSG boundary, so whole features can spill past it. Clip them to the
# boundary shape so nothing renders outside the basin.
clip_aoi <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(x)
  out <- suppressWarnings(sf::st_intersection(x, sf::st_union(sf::st_geometry(aoi))))
  if (nrow(out) == 0L) NULL else out
}
roads    <- clip_aoi(roads)
railways <- clip_aoi(railways)
reserves <- clip_aoi(reserves)
parks    <- clip_aoi(parks)
```

## Reproducing bcfishpass (parity) {-}

We compare our per-segment classification against the `bcfishpass` snapshot,
segment by segment. Of the two species, only bull trout occurs in the watershed
group, so this is a single-species comparison.

```{r link-parity-table}
ptab <- parity[, c("species", "total_segs", "match_pct", "n_diffs")]
names(ptab) <- c("species", "segments", "match %", "n diffs")
knitr::kable(
  ptab, row.names = FALSE,
  caption = "Per-segment classification parity for bull trout, our values vs the bcfishpass snapshot."
)
```

```{r link-parity-pct, results = "asis"}
cat(sprintf(paste0("We reproduce **%.2f%%** of bcfishpass's per-segment ",
                   "bull-trout classification across %s segments, with %s ",
                   "disagreements, consistent with the ~99.7%% study-area ",
                   "median we see across the Peace.\n"),
            parity$match_pct[1],
            format(parity$total_segs[1], big.mark = ","),
            format(parity$n_diffs[1], big.mark = ",")))
```

```{r link-symbology, include = FALSE}
reg <- gq::gq_reg_main()
cls <- gq::gq_tmap_classes(reg$layers$streams_salmon)

# bcfishpass line widths come from the same registry — habitat use drives line
# weight (SPAWN thick, REAR medium, ACCESS thin), barrier status drives colour.
# Stored raw here and scaled per map below (full-WSG views need thinner lines
# than the zoomed detail panels). Unmatched tokens fall to the thinnest weight.
width_for <- function(tokens) {
  w <- cls$widths[tokens]
  w[is.na(w)] <- min(cls$widths, na.rm = TRUE)
  unname(w)
}

# Legend labels are decoded from the mapping_code token rather than taken from
# cls$labels: the upstream QGIS project mislabels the ACCESS habitat token as
# "No known barriers" (colliding with token2's own status wording, producing
# "No known barriers; known barrier"). Fix-at-source tracked in the bcfishpass
# fork and gq registry; decoded here so the legend reads correctly meanwhile.
.tok1 <- c(SPAWN = "Spawning", REAR = "Rearing", ACCESS = "Accessible")
.tok2 <- c(NONE = "no known barriers", MODELLED = "potential barrier",
           ASSESSED = "known barrier", DAM = "dam", REMEDIATED = "remediated")
label_for <- function(tokens) {
  vapply(strsplit(tokens, ";", fixed = TRUE), function(p) {
    parts <- c(.tok1[p[1]], if (length(p) >= 2) .tok2[p[2]],
               if (length(p) >= 3 && p[3] == "INTERMITTENT") "intermittent")
    paste(parts[!is.na(parts)], collapse = "; ")
  }, character(1))
}

# Context layers shared by the two full-WSG maps, drawn UNDER the
# mapping_code streams: waterbodies, resource roads, railways, then the
# land-use polygons (parks, reserves) on top of hydro/transport so they read
# where they overlap — the flooded layering. Reserves are tiny at WSG scale,
# so each gets a centroid diamond marker.
draw_context_full <- function() {
  if (!is.null(waterbodies)) {
    plot(sf::st_geometry(waterbodies), col = "#a3cdb966", border = "#2171B5",
         lwd = 0.3, add = TRUE)
  }
  if (!is.null(roads)) {
    plot(sf::st_geometry(roads), col = "#88888888", lwd = 0.25, add = TRUE)
  }
  if (!is.null(railways)) {
    plot(sf::st_geometry(railways), col = "black", lwd = 0.8, lty = "dashed",
         add = TRUE)
  }
  if (!is.null(parks)) {
    plot(sf::st_geometry(parks), col = "#639b5f55", border = "#33a02c",
         lwd = 0.7, add = TRUE)
  }
}

# Reserves draw OVER the mapping_code streams (called after the stream plot),
# so the polygon + centroid diamond read where they overlap the network.
draw_reserves <- function(x, fill = "#b2b2b288", cex = 0.9) {
  if (is.null(x) || nrow(x) == 0L) return(invisible())
  plot(sf::st_geometry(x), col = fill, border = "#232323", lwd = 0.6,
       add = TRUE)
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(x),
                                          of_largest_polygon = TRUE))
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = cex, add = TRUE)
}

# Reserve english_name labels with a thin white halo, drawn last with
# xpd = TRUE so an edge label isn't cropped at the plot region.
label_reserves <- function(x, cex = 0.45) {
  if (is.null(x) || !"english_name" %in% names(x) || nrow(x) == 0L) {
    return(invisible())
  }
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(x),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  ofs_y  <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) {
    text(coords[, 1] + dx * halo_x, ly, labels = x$english_name,
         cex = cex, col = "white", font = 2, xpd = TRUE)
  }
  for (dy in c(-1, 1)) {
    text(coords[, 1], ly + dy * halo_y, labels = x$english_name,
         cex = cex, col = "white", font = 2, xpd = TRUE)
  }
  text(coords[, 1], ly, labels = x$english_name, cex = cex, col = "#111111",
       font = 2, xpd = TRUE)
}

# mapping_code legend drawn outside the plotting region (into the reserved
# right margin) so it never overlaps the network.
legend_mapping_code <- function(tokens) {
  present <- names(cls$values) %in% unique(tokens)
  toks <- names(cls$values)[present]
  key_lwd <- cls$widths[present] * 2
  # base R scales dash length by line width, so a thick "dashed" key fits only
  # ~1 dash in the short legend segment while thin keys show many. Give thicker
  # intermittent keys a finer pattern so all read as dashes at the same seg.len.
  is_int <- grepl(";INTERMITTENT", toks)
  key_lty <- ifelse(!is_int, "solid",
                    ifelse(key_lwd > 2.5, "22",
                           ifelse(key_lwd > 1.5, "33", "44")))
  legend("topright", inset = c(-0.34, 0), xpd = TRUE, title = "Legend",
         legend = label_for(toks), col = cls$values[present],
         lwd = key_lwd, lty = key_lty,
         seg.len = 3, cex = 0.55, bty = "n")
}
```

```{r link-map-bt, fig.cap = "Bull-trout habitat and connectivity classification across the Parsnip River Watershed Group. Each stream segment is coloured by the most significant barrier downstream and weighted by habitat value; the legend gives the full key."}
bt <- streams[!is.na(streams$mapping_code_bt) & nzchar(streams$mapping_code_bt), ]
bt$col <- cls$values[bt$mapping_code_bt]
bt$col[is.na(bt$col)] <- "#999999"
bt$w <- width_for(bt$mapping_code_bt)
bt$lty <- ifelse(grepl(";INTERMITTENT", bt$mapping_code_bt), "dashed", "solid")

op <- par(mar = c(2, 1, 4, 8))
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, axes = FALSE,
     main = "Bull trout — Parsnip River Watershed Group")
draw_context_full()
plot(sf::st_geometry(bt), col = bt$col, lwd = bt$w * 0.6, lty = bt$lty, add = TRUE)
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, add = TRUE)
draw_reserves(reserves)
label_reserves(reserves)
legend_mapping_code(bt$mapping_code_bt)
par(op)
```

## Arctic grayling — extending the method {-}

`bcfishpass` does not yet model Arctic grayling, so this is net-new output —
there is nothing to compare against. We apply the same method, re-parameterised
for grayling, and map the result with the same symbology so the two species are
directly comparable.

```{r link-map-gr, fig.cap = "Arctic grayling habitat and connectivity classification across the Parsnip River Watershed Group — a species bcfishpass does not yet model. Same symbology as the bull-trout map, so the two are directly comparable; grayling's modelled network is the smaller of the two."}
gr <- streams[!is.na(streams$mapping_code_gr) & nzchar(streams$mapping_code_gr), ]
gr$col <- cls$values[gr$mapping_code_gr]
gr$col[is.na(gr$col)] <- "#999999"
gr$w <- width_for(gr$mapping_code_gr)
gr$lty <- ifelse(grepl(";INTERMITTENT", gr$mapping_code_gr), "dashed", "solid")

op <- par(mar = c(2, 1, 4, 8))
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, axes = FALSE,
     main = "Arctic grayling — Parsnip River Watershed Group")
draw_context_full()
plot(sf::st_geometry(gr), col = gr$col, lwd = gr$w * 0.6, lty = gr$lty, add = TRUE)
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, add = TRUE)
draw_reserves(reserves)
label_reserves(reserves)
legend_mapping_code(gr$mapping_code_gr)
par(op)
```

## Maps — detail comparison {-}

The full-watershed views compress a lot of network. Cropping to a sub-reach puts
bull trout and grayling side by side at full resolution — the reaches where
grayling carries a classification that bull trout does not are where the
extension does genuinely new work.

```{r link-map-detail, fig.width = 7, fig.height = 4.5, fig.cap = "South-east corner of the Parsnip River Watershed Group at full resolution: bull trout (left) and Arctic grayling (right), same extent and symbology. Grey background streams are the full modelled network, so the coloured overlay shows where each species' classification reaches."}
e <- sf::st_bbox(aoi)
inset_bbox <- sf::st_bbox(c(
  xmin = unname(e["xmin"] + (e["xmax"] - e["xmin"]) * 0.55),
  ymin = unname(e["ymin"]),
  xmax = unname(e["xmax"]),
  ymax = unname(e["ymin"] + (e["ymax"] - e["ymin"]) * 0.45)
), crs = sf::st_crs(aoi))

crop_sf <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(NULL)
  out <- suppressWarnings(sf::st_crop(x, inset_bbox))
  if (nrow(out) == 0L) NULL else out
}

frame <- sf::st_as_sfc(inset_bbox)
all_d   <- crop_sf(streams)
bt_d    <- crop_sf(bt)
gr_d    <- crop_sf(gr)
wb_d    <- crop_sf(waterbodies)
aoi_d   <- crop_sf(aoi)
roads_d <- crop_sf(roads)
rail_d  <- crop_sf(railways)
res_d   <- crop_sf(reserves)
parks_d <- crop_sf(parks)
ns_d    <- crop_sf(named_streams)

panel <- function(seg, title) {
  plot(sf::st_geometry(frame), col = NA, border = NA, axes = FALSE,
       main = title)
  if (!is.null(wb_d)) {
    plot(sf::st_geometry(wb_d), col = "#a3cdb988", border = "#2171B5",
         lwd = 0.4, add = TRUE)
  }
  if (!is.null(all_d)) {
    plot(sf::st_geometry(all_d), col = "#cccccc", lwd = 0.4, add = TRUE)
  }
  if (!is.null(roads_d)) {
    plot(sf::st_geometry(roads_d), col = "#666666aa", lwd = 0.4, add = TRUE)
  }
  if (!is.null(rail_d)) {
    plot(sf::st_geometry(rail_d), col = "black", lwd = 1.0, lty = "dashed",
         add = TRUE)
  }
  if (!is.null(parks_d)) {
    plot(sf::st_geometry(parks_d), col = "#639b5f55", border = "#33a02c",
         lwd = 0.8, add = TRUE)
  }
  if (!is.null(seg) && nrow(seg) > 0L) {
    plot(sf::st_geometry(seg), col = seg$col, lwd = seg$w, lty = seg$lty,
         add = TRUE)
  }
  draw_reserves(res_d, fill = "#b2b2b2aa", cex = 1.0)
  if (!is.null(aoi_d)) {
    plot(sf::st_geometry(aoi_d), border = "black", lwd = 1.5, add = TRUE)
  }
  if (!is.null(ns_d)) {
    dedup <- ns_d[!duplicated(ns_d$gnis_name), ]
    npts <- suppressWarnings(sf::st_centroid(sf::st_geometry(dedup)))
    text(sf::st_coordinates(npts), labels = dedup$gnis_name, cex = 0.45,
         font = 3, col = "#0d3a6c")
  }
}

op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
panel(bt_d, "Bull trout")
panel(gr_d, "Arctic grayling")
par(op)
```
