---
output: html_document
editor_options:
  chunk_output_type: console
---


# Appendix - Assessment Data Summary {-#app-site-data}

Fish passage assessment procedures conducted through SERNbc since `r params$project_start_year` are amalgamated in Tables \@ref(tab:tab-sites-sum-cap) - \@ref(tab:tab-sites-cap). Since `r params$project_start_year`, orthoimagery and elevation model rasters have been generated and stored as Cloud Optimized Geotiffs on a cloud service provider (AWS) with select imagery linked to in the collaborative GIS project. Additionally, a tile service has been set up to facilitate viewing and downloading of individual images, provided in [Appendix - UAV Imagery](#app-uav).

`r if(gitbook_on){knitr::asis_output("<br>")}`

```{r ld-db}
sites_all <- arrow::read_parquet("data/snapshots/fp_sites_tracking.parquet")

```

```{r summarize-sites}

wsg <- c(
  "Lower Chilako River",
  "Willow River",
  "Tabor River",
  "Lower Salmon River",
  "Upper Fraser River",
  "Nechako River",
  "Morkill River",
  "Francois Lake"
)

sites_all_summary <- sites_all |>
  dplyr::mutate(
    uav = dplyr::case_when(
      !is.na(link_uav1) ~ "yes",
      T ~ NA_character_
    )) |>
  dplyr::filter(watershed_group %in% wsg) |>
  dplyr::group_by(watershed_group) |>
  dplyr::summarise(
    dplyr::across(assessment:fish_sampling, ~ sum(!is.na(.x))),
    uav = sum(!is.na(uav))
  ) |>
  sf::st_drop_geometry() |>
  dplyr::rename_with(~ stringr::str_replace_all(., "_", " ") |>
                       stringr::str_to_title()) |>
  dplyr::rename(
    `Drone Imagery` = Uav) |>
  janitor::adorn_totals()
```

```{r tab-sites-sum-cap, results="asis"}
my_caption = "Summary of fish passage assessment procedures conducted in northern British Columbia through SERNbc."
my_tab_caption()
```

```{r tab-sites-sum}
sites_all_summary |>
  dplyr::mutate(dplyr::across(everything(), as.character)) |>
  my_dt_table(
    page_length = 20,
    cols_freeze_left = 0
              )
```

<br>


```{r tab-sites-cap, results="asis"}
my_caption = "Details of fish passage assessment procedures conducted in northern British Columbia through SERNbc."
my_tab_caption(tip_flag = FALSE)
```

```{r tab-sites-all}
sites_all |>
  dplyr::filter(watershed_group %in% wsg) |>
  sf::st_drop_geometry() |>
  dplyr::mutate(dplyr::across(everything(), as.character)) |>
  dplyr::relocate(watershed_group, .after = my_crossing_reference) |>
  dplyr::select(-idx) |>
  dplyr::rename_with(~ . |>
                       stringr::str_replace_all("_", " ") |>
                       stringr::str_replace_all("repo", "Report") |>
                       stringr::str_replace_all("uav", "Drone") |>
                       stringr::str_to_title()) |>
  my_dt_table(
    cols_freeze_left = 1,
    escape = FALSE
  )
```
