---
output: html_document
editor_options:
  chunk_output_type: console
---


# **Appendix - Environmental DNA Results** {-#app-edna}

This appendix contains site-level detail for the 2025 environmental DNA (eDNA) sampling program in the Peace project area. Headline per-species detection counts are presented in the main Results chapter; the tables below show which sites detected which species. Field blanks are tracked separately as quality-assurance samples — they test for protocol contamination, not site eDNA. Office blanks (samples filtered at the field accommodation rather than at site) are excluded from these tables because their UTM coordinates are inherited from related sites and do not represent real sampling locations.

The interactive map of these results is linked from the main Results chapter and at the end of this appendix.

```{r app-edna-prep, include=FALSE}
# Loads the snapshotted by_site_target CSV and prepares per-site /
# field-blank / retest views used by the tables in this appendix.
# (Same pattern as 0400-results.Rmd's prep chunk; both could be lifted into
# a shared helper once a third consumer needs it.)

norm_lgl <- function(x) {
  if (is.logical(x)) return(x)
  out <- toupper(as.character(x)) == "TRUE"
  out[is.na(x)] <- FALSE
  out
}

# `edna_species_names` and `fmt_targets` are defined globally in 0400-results.Rmd's
# Phase 2 prep chunk. They flow through to this appendix via the bookdown merge
# order (0400 < 0850 alphabetically). Don't redefine here — the prior local copy
# drifted from main Results' Kokanee labelling and produced inconsistent species
# names in the same report (caught by /code-check 2026-05-05).

edna_bystargets_app <- readr::read_csv(
  "data/edna_unbc_results_2025_by_site_target.csv",
  show_col_types = FALSE
) |>
  dplyr::filter(grepl("sern_peace_fwcp_2023", source)) |>
  dplyr::mutate(
    control_blank_field           = norm_lgl(control_blank_field),
    control_blank_office          = norm_lgl(control_blank_office),
    control_species_present_field = norm_lgl(control_species_present_field)
  ) |>
  dplyr::filter(!control_blank_office)  # office blanks dropped (fake coords)

# --- Per-site rollup (real environmental samples only) ---------------------

edna_real_app <- edna_bystargets_app |>
  dplyr::filter(!control_blank_field)

edna_per_site <- edna_real_app |>
  dplyr::group_by(site_id, stream_name) |>
  dplyr::summarise(
    `UNBC ID`         = paste(sort(unique(unlist(strsplit(lab_ids, ", ")))), collapse = ", "),
    `Pos. control`    = ifelse(any(control_species_present_field), "yes", ""),
    `Detected`        = fmt_targets(target[any_clean],
                                    max_pos[any_clean],
                                    retested[any_clean]),
    `Sub-threshold`   = fmt_targets(target[any_crude & !any_clean],
                                    max_pos[any_crude & !any_clean],
                                    retested[any_crude & !any_clean]),
    `Not detected`    = fmt_targets(target[!any_crude],
                                    max_pos[!any_crude],
                                    retested[!any_crude]),
    .groups = "drop"
  ) |>
  dplyr::rename(`Site ID` = site_id, `Stream` = stream_name) |>
  dplyr::arrange(`Site ID`)

# --- Field blanks (separate table) -----------------------------------------

edna_field_blanks <- edna_bystargets_app |>
  dplyr::filter(control_blank_field) |>
  dplyr::group_by(site_id, stream_name) |>
  dplyr::summarise(
    `UNBC ID`         = paste(sort(unique(unlist(strsplit(lab_ids, ", ")))), collapse = ", "),
    `Detected`        = fmt_targets(target[any_clean],
                                    max_pos[any_clean],
                                    retested[any_clean]),
    `Sub-threshold`   = fmt_targets(target[any_crude & !any_clean],
                                    max_pos[any_crude & !any_clean],
                                    retested[any_crude & !any_clean]),
    `Clean (0 droplets)` = fmt_targets(target[!any_crude],
                                       max_pos[!any_crude],
                                       retested[!any_crude]),
    .groups = "drop"
  ) |>
  dplyr::rename(`Site ID` = site_id, `Stream` = stream_name) |>
  dplyr::arrange(`Site ID`)

# --- Retests (sites x targets that UNBC re-ran) ----------------------------

edna_retests <- edna_bystargets_app |>
  dplyr::filter(retested) |>
  dplyr::mutate(
    Species   = dplyr::coalesce(edna_species_names[target], target),
    `Result`  = dplyr::case_when(
      any_clean             ~ "Detected (≥4 droplets)",
      any_crude             ~ "Sub-threshold (1-3 droplets, NOT a detection)",
      TRUE                  ~ "No detection (0 droplets)"
    ),
    `Sample type` = dplyr::case_when(
      control_blank_field ~ "field blank",
      TRUE                ~ "environmental"
    )
  ) |>
  dplyr::select(
    `Site ID`        = site_id,
    Stream           = stream_name,
    Species,
    `Code`           = target,
    `Sample type`,
    `Runs`           = n_runs,
    `Max droplets`   = max_pos,
    Result
  ) |>
  dplyr::arrange(`Site ID`, Code)
```

## Per-site detection results {-}

Format: `CODE(max_droplets)`, with `*` flagging samples that UNBC reran. Detected = ≥4 positive droplets per the standard ddPCR call threshold; Sub-threshold = 1-3 droplets, not interpretable as species presence per ddPCR convention; Not detected = 0 droplets across all runs.

```{r tab-edna-per-site}
edna_per_site |>
  fpr::fpr_kable(
    caption_text = "Site-level eDNA detection results across real environmental samples in the Peace project area. UNBC ID is the lab sample identifier; Pos. control = TRUE indicates the sample was collected at a site where the target species was also confirmed by electrofishing.",
    scroll = gitbook_on
  )
```

<br>

## Field blanks {-}

Field blanks filter distilled water at the field site to test for protocol contamination during collection and processing. **Detections in blanks indicate potential cross-contamination, not species presence at the location** — these rows are presented for QA transparency only, not as eDNA detections.

```{r tab-edna-field-blanks}
edna_field_blanks |>
  fpr::fpr_kable(
    caption_text = "Field blanks collected during the 2025 eDNA program in the Peace project area. Detections in blanks reflect protocol-side contamination, not site eDNA.",
    scroll = FALSE
  )
```

<br>

## Retests {-}

UNBC reran a subset of samples to confirm borderline results. The table below lists every (site × species) combination that was rerun, including reruns from both real environmental samples and field blanks.

```{r tab-edna-retests, eval = TRUE}
if (nrow(edna_retests) > 0) {
  edna_retests |>
    fpr::fpr_kable(
      caption_text = "UNBC retests for the 2025 Peace eDNA samples. Runs is the total number of times the sample x species combination was assayed; Max droplets is the highest positive-droplet count observed across all runs.",
      scroll = gitbook_on
    )
} else {
  cat("No retests in the Peace 2025 batch.\n")
}
```

<br>

## Interactive map {-}

The interactive Peace eDNA map shows site-level detection results with toggleable per-species layers, an opt-in sub-threshold-signals layer (off by default), and an opt-in Controls layer for field blanks. See the `r ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "edna_unbc_results_2025_peace_map.html", anchor_text = "interactive Peace eDNA map")`.
