---
output: html_document
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup-198692-mon, eval = F}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
source('scripts/packages.R')
source('scripts/tables.R')
source('scripts/functions.R')
```

```{r load-data-198692-mon}
my_site <- 198692
```

# Tributary to Kerry Lake - `r as.character(my_site)` - Monitoring Appendix {- #trib-to-kerry-2025}

Baseline monitoring was conducted in `r params$project_year` at PSCIS crossing `r as.character(my_site)` ahead of planned remediation, to inform future effectiveness monitoring once the structure is upgraded. Site location, background, and the 2024 habitat confirmation assessment are documented in @irvine_schick2025RestoringFish; the 2024 site appendix is available at [https://www.newgraphenvironment.com/fish_passage_peace_2024_reporting/trib-to-kerry.html](https://www.newgraphenvironment.com/fish_passage_peace_2024_reporting/trib-to-kerry.html).

<br>

## Fish Sampling {.unnumbered}

Electrofishing was conducted downstream and upstream of PSCIS crossing `r as.character(my_site)` in `r params$project_year` to establish a pre-remediation baseline. Per-site capture results are in Table \@ref(tab:tab-fish-site-198692-mon); densities are in Table \@ref(tab:tab-fish-dens-198692-mon).

<br>

```{r tab-fish-site-198692-mon, eval = T}
fpr::fpr_table_fish_site()
```

<br>

```{r tab-fish-dens-198692-mon, eval = T}
fpr::fpr_table_fish_density()
```

<br>

## Environmental DNA {.unnumbered}

eDNA samples were collected upstream and downstream of PSCIS crossing `r as.character(my_site)` in `r params$project_year` as part of the program-wide eDNA sampling effort. Per-position detection results are presented in Table \@ref(tab:tab-edna-results-198692-mon). Detections are based on the standard ddPCR call threshold of ≥4 positive droplets; format is `CODE(max_droplets)`, with `*` flagging samples UNBC reran. Site-level detail across all sampled crossings is in [Appendix - Environmental DNA Results](#app-edna), and the `r ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "edna_unbc_results_2025_peace_map.html", anchor_text = "interactive Peace eDNA map")` provides a toggleable per-species view.

<br>

```{r tab-edna-results-198692-mon-prep, include=FALSE}
edna_results_198692_mon <- edna_bystargets_peace |>
  dplyr::filter(grepl(paste0("^", my_site), site_id), !control_blank_field, !control_blank_office) |>
  dplyr::group_by(site_id) |>
  dplyr::summarise(
    `UNBC ID`        = paste(sort(unique(unlist(strsplit(lab_ids, ", ")))), collapse = ", "),
    `Detected`       = fmt_targets(target[any_clean], max_pos[any_clean], retested[any_clean]),
    `Sub-threshold`  = fmt_targets(target[any_crude & !any_clean],
                                   max_pos[any_crude & !any_clean],
                                   retested[any_crude & !any_clean]),
    `Not detected`   = fmt_targets(target[!any_crude],
                                   max_pos[!any_crude],
                                   retested[!any_crude]),
    .groups = "drop"
  ) |>
  dplyr::rename(`Site ID` = site_id) |>
  dplyr::arrange(`Site ID`)
```

```{r tab-edna-results-198692-mon}
edna_results_198692_mon |>
  fpr::fpr_kable(
    caption_text = paste0("eDNA detection results at crossing ", my_site, "."),
    scroll = FALSE
  )
```

<br>

## Conclusion {.unnumbered}

Baseline fish and eDNA data were collected at PSCIS crossing `r as.character(my_site)` in `r params$project_year` to inform future effectiveness monitoring once remediation completes. Resampling at the site once the structure is upgraded will allow direct comparison of fish abundance and habitat conditions before and after the intervention.
