---
output: html_document
editor_options:
  chunk_output_type: console
---

# **Appendix - Climate Departure** {-#app-climate-departure}

```{r cd-setup, include = FALSE}
library(cd)
library(sf)
library(terra)
library(ggplot2)
library(tidyterra)
```

```{r cd-load}
gpkg_cd    <- "data/gis/cd_peace.gpkg"
aoi          <- sf::st_read(gpkg_cd, layer = "aoi",        quiet = TRUE)
ecoregions   <- sf::st_read(gpkg_cd, layer = "ecoregions", quiet = TRUE)
ecoregions   <- ecoregions[ecoregions$area_km2 > 100, ]
ecoregions$name_tc <- tools::toTitleCase(tolower(ecoregions$name))
towns        <- sf::st_read(gpkg_cd, layer = "towns",      quiet = TRUE)
lakes        <- sf::st_read(gpkg_cd, layer = "lakes",      quiet = TRUE)
rivers       <- sf::st_read(gpkg_cd, layer = "rivers",     quiet = TRUE)
streams      <- sf::st_read(gpkg_cd, layer = "streams",    quiet = TRUE)
highways     <- sf::st_read(gpkg_cd, layer = "highways",   quiet = TRUE)
wsgs         <- sf::st_read(gpkg_cd, layer = "wsgs",       quiet = TRUE)

cd_data    <- readRDS("data/gis/cd_peace.rds")
ts         <- cd_data$regional$ts
bl_early   <- cd_data$regional$bl
ano        <- cd_data$regional$ano
trn        <- cd_data$regional$trn
cmp        <- cd_data$regional$cmp
cmp_pct    <- cd_data$regional$cmp_pct
departure  <- terra::rast("data/gis/cd_peace_departure_tmean.tif")
names(departure) <- "Temperature departure"

aoi_km2 <- as.numeric(sum(sf::st_area(sf::st_transform(aoi, 3005)))) / 1e6
aoi_bb  <- sf::st_bbox(aoi)
```

## Climate departure and fish passage {-}

Stream crossings are the visible structures in a fish-passage
inventory: culverts, bridges, weirs, fords. The streams they cross
sit inside a climate that is changing — warmer air, shifting
snowpack, longer shoulder seasons — and those climate changes set
the hydrologic and thermal context the structures pass fish
*through*. A culvert that worked when it was installed in 1985 is
now passing a different hydrograph: earlier freshet, lower
late-summer flows, warmer water. The barrier inventory does not see
that on its own.

Warmer air drives warmer streams. For the cold-water salmonids the
Peace supports — bull trout, Arctic grayling, mountain whitefish —
that can move habitat in either direction: cold-limited headwaters
gain growing-degree-days (larger fish, better overwinter survival,
stronger out-migration), while reaches near the upper edge of the
thermal niche lose usable habitat. The recovery value of removing a
downstream barrier scales with which side of that line each
upstream reach sits on. Climate departure tells you which side that
is, watershed by watershed.

Beyond temperature, hydrologic timing matters. Earlier and faster
snowmelt shifts the freshet weeks earlier in the calendar and
compresses the high-flow pulse. That changes design assumptions for
crossing structures, and it changes the timing of migration, gravel
mobilisation, and off-channel rearing — the in-stream processes
barrier removal is meant to restore.

Quantifying departure at the watershed scale lets us carry climate
context into barrier prioritisation alongside the structural and
biological inputs.

## Area of interest {-}

The FWCP Peace Region is a `r format(round(aoi_km2), big.mark = ",")`
km² administrative boundary in northeastern British Columbia covering
Williston Reservoir and surrounding watersheds. All watersheds drain
to the Mackenzie River — the Peace flows east into the Slave, which
flows into the Mackenzie, which empties into the Beaufort Sea on the
Arctic Ocean. The region spans `r nrow(ecoregions)` ecoregions
(Figure \@ref(fig:cd-map-aoi)) — broad climate-physiography zones
based on latitude, elevation, and dominant vegetation — and we
carry that ecoregion partition through the analysis because
climate departure can vary across them in ways the regional
average hides.

```{r cd-map-aoi, fig.cap = "FWCP Peace Region area of interest in northeastern British Columbia, coloured by ecoregion. Williston Reservoir dominates the basin; the climate-departure pipeline is run on the regional outline (heavy blue) and again on each ecoregion polygon.", fig.height = 6.5}
ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi,      fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = lakes,    fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = rivers,   fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = streams,  color = "#74a9cf", linewidth = 0.3, alpha = 0.6) +
  geom_sf(data = highways, color = "#9c9c9c", linewidth = 0.75) +
  geom_sf(data = highways, color = "#ffe585", linewidth = 0.5) +
  geom_sf(data = towns,    color = "black", size = 2.5) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3.2, fill = "white", alpha = 0.85,
    label.padding = unit(0.2, "lines"), min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
           ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)) +
  labs(title = "FWCP Peace Region",
       subtitle = paste0(format(round(aoi_km2), big.mark = ","),
                         " km² — ", nrow(ecoregions), " ecoregions")) +
  theme_minimal(base_size = 11) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text  = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

## Recent decade versus pre-warming reference {-}

The table below compares two windows directly — the recent decade
(2015–2025) against the pre-warming reference (1951–1980). The
difference is reported in the variable's native units, alongside
percent change where it is meaningful (precipitation, soil
moisture, vapour pressure deficit, relative humidity). Two
p-values appear, answering different questions. **Δ p (windows)**
is the Welch t-test on the annual values within each window — this
tells us if the recent decade differs from the pre-warming
reference. **Trend p (75-yr)** is the Mann-Kendall test on the full
1951–present series — this tells us if there is a steady
year-on-year ramp. Step changes show up as significant Δ p with
non-significant trend p; gradual ramps show up as both significant.

```{r cd-compare-table}
trn_p <- trn[trn$trend_start == 1951, c("variable", "period", "mk_pvalue")]
names(trn_p)[3] <- "trend_p"
trn_p$trend_p <- round(trn_p$trend_p, 3)

no_pct_vars <- c("tmean", "tmax", "tmin",
                 "snow_cover", "snowfall_fraction", "snowmelt_doy_50")
cmp_combined <- cmp
cmp_combined$pct_change <- ifelse(
  cmp_combined$variable %in% no_pct_vars,
  NA_real_,
  round(100 * (cmp_combined$mean_a - cmp_combined$mean_b) / cmp_combined$mean_b, 1)
)
cmp_combined$mean_a   <- round(cmp_combined$mean_a, 2)
cmp_combined$mean_b   <- round(cmp_combined$mean_b, 2)
cmp_combined$abs_diff <- round(cmp_combined$difference, 2)
cmp_live <- cd::cd_compare(ts)
cmp_combined$window_p <- round(
  cmp_live$p_value[match(paste(cmp_combined$variable, cmp_combined$period),
                          paste(cmp_live$variable, cmp_live$period))], 3)

cmp_combined <- cmp_combined[, c("variable", "period", "mean_a", "mean_b",
                                  "abs_diff", "pct_change", "window_p")]
cmp_combined <- merge(cmp_combined, trn_p,
                      by = c("variable", "period"), all.x = TRUE)
cmp_combined <- cmp_combined[order(cmp_combined$variable, cmp_combined$period), ]
names(cmp_combined) <- c("Variable", "Period",
                         "Recent (2015–2025)", "Pre-warming (1951–1980)",
                         "Δ absolute", "Δ %",
                         "Δ p (windows)", "Trend p (75-yr)")

kableExtra::kable_styling(
  knitr::kable(cmp_combined, label = NA, row.names = FALSE,
    caption = "Recent decade (2015-2025) versus pre-warming reference (1951-1980) for all 15 climate variables and all available periods, FWCP Peace Region. Two p-values answer different questions — see text."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "420px")
```

<br>

The recent decade ran
`r sprintf("%.1f", cmp$difference[cmp$variable == "tmean" & cmp$period == "annual"])` °C
warmer than the pre-warming reference for annual mean temperature,
with both window and trend p-values below 0.001. Vapour pressure
deficit rose significantly on both tests. Annual precipitation rose
about 4 % but neither test confirms the shift. Soil moisture and
relative humidity show no meaningful change. The snow rows
(swe summer, snowmelt spring, snowmelt_doy_50) carry the headline
snowpack-timing signals discussed below.

## Trends over the long record {-}

Anomalies are computed against the 1951–1980 pre-warming reference.
Saying a year is "+1.5 °C" means it was 1.5 °C warmer than the
average year between 1951 and 1980. The trend table below has two
rows per variable — trends computed from two different start years:

- **1951–present (75 years)** — the long view. Captures the full
  magnitude of warming since the pre-warming reference.
- **1981–present (45 years)** — starts at the beginning of the
  current 30-year climate normal (1981–2010), the reference used in
  most published climate products.

Comparing the two slopes is informative. If the 45-year slope is
steeper than the 75-year slope, warming has accelerated; if
shallower, warming has slowed (though "slower" almost never means
"stopped"). Similar slopes mean a roughly steady rate of change
across the full record. Total change is slope multiplied by the
number of years — the cumulative shift over the trend window.

```{r cd-trend-table}
kableExtra::kable_styling(
  knitr::kable(cd::cd_summary(trn), label = NA,
    caption = "Trend statistics for all variables and periods, FWCP Peace Region — Theil-Sen slope, Mann-Kendall p-value, and cumulative change over each trend window."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

```{r cd-plot-tmean, fig.cap = "Annual mean temperature anomaly for the FWCP Peace Region relative to the 1951-1980 baseline. Bars are degrees above (red) or below (blue) the pre-warming average; the two trend lines are the 75-year (1951-present, dashed) and 45-year (1981-present, solid) Theil-Sen slopes."}
cd::cd_plot_timeseries(
  ano, variable = "tmean", period = "annual", trend = trn,
  title = "Annual Mean Temperature Anomaly — FWCP Peace Region"
)
```

```{r cd-plot-prcp, fig.cap = "Annual precipitation anomaly (% of the 1951-1980 baseline) for the FWCP Peace Region. The long-term trend is positive but not statistically significant at the regional scale; significance does appear in the two northernmost ecoregions (see Per-Ecoregion section)."}
cd::cd_plot_timeseries(
  ano, variable = "prcp", period = "annual", trend = trn,
  title = "Annual Precipitation Anomaly — FWCP Peace Region"
)
```

## Daytime highs and overnight lows {-}

The `cd` package ships daytime maximum (tmax) and overnight minimum
(tmin) temperatures alongside the daily mean. They carry distinct
information. Overnight minimums warming faster than daytime
maximums — the "day-night asymmetry" — is one of the textbook
fingerprints of greenhouse warming. @karl_etal1993NewPerspective
documented this first at the global scale, finding overnight
minimums rose at roughly three times the rate of daytime maximums
between 1951 and 1990.

For the FWCP Peace Region, overnight minimums are warming faster
than daytime maximums. Daytime maximums warmed about +0.027 °C per
year since 1951 (+2.0 °C cumulative); overnight minimums warmed
about +0.032 °C per year (+2.4 °C cumulative). The overnight side
warmed roughly 0.4 °C more than the daytime side over the full
record, narrowing the diurnal temperature range by the same amount
(Figure \@ref(fig:cd-plot-dtr)).

```{r cd-plot-tmax, fig.cap = "Annual daytime maximum temperature (tmax) anomaly relative to the 1951-1980 baseline."}
trn_tmax <- cd::cd_trend(
  ano[ano$variable == "tmax" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd::cd_plot_timeseries(ano, variable = "tmax", period = "annual", trend = trn_tmax,
  title = "Daytime Maximum (tmax) — Annual Anomaly")
```

```{r cd-plot-tmin, fig.cap = "Annual overnight minimum temperature (tmin) anomaly relative to the 1951-1980 baseline."}
trn_tmin <- cd::cd_trend(
  ano[ano$variable == "tmin" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd::cd_plot_timeseries(ano, variable = "tmin", period = "annual", trend = trn_tmin,
  title = "Overnight Minimum (tmin) — Annual Anomaly")
```

```{r cd-plot-dtr, fig.cap = "Diurnal temperature range (daytime maximum minus overnight minimum) annual mean for the FWCP Peace Region. The downward slope is the day-night asymmetry — overnight lows are warming faster than daytime highs."}
tmax_ts <- ts[ts$variable == "tmax" & ts$period == "annual", c("year", "value")]
tmin_ts <- ts[ts$variable == "tmin" & ts$period == "annual", c("year", "value")]
dtr <- merge(tmax_ts, tmin_ts, by = "year", suffixes = c("_max", "_min"))
dtr$dtr <- dtr$value_max - dtr$value_min

ggplot(dtr, aes(x = year, y = dtr)) +
  geom_line(color = "grey50") +
  geom_point(color = "grey30", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#b2182b", linewidth = 0.8) +
  labs(title = "Diurnal Temperature Range — FWCP Peace Region",
       x = NULL,
       y = expression("Daytime maximum minus overnight minimum (" * degree * "C)")) +
  theme_minimal(base_size = 11)
```

## Snowpack {-}

Snowpack drives the seasonal water supply across British Columbia:
winter precipitation falls as snow, accumulates on the ground, and
releases as meltwater across spring and summer. That seasonal
storage is the difference between a late-summer creek that still
flows and one that does not. It also sets the timing of the spring
freshet — the annual flood pulse that salmonids time their
up-river migrations to.

For the FWCP Peace Region the snowpack signal is unambiguous and
concentrated in the warm shoulders of the year. Annual snow water
equivalent declined about 10 % (135 → 122 mm), but the seasonal
breakdown is sharper: **summer snow water equivalent collapsed by 75 %** (21.5 →
5.3 mm) and **spring snowmelt rose 37 %** (212 → 290 mm) as the
freshet shifted earlier in the calendar. Total annual snowfall
barely changed (-6 %) — the snow water equivalent decline is mostly about warming
removing snow earlier, not less snow falling.

### Seasonal snowpack curve {-}

Aggregated to the four standard meteorological seasons — winter
(December–February), spring (March–May), summer (June–August), and
fall (September–November) — the table below compares the recent
decade (2015–2025) directly against the pre-warming reference
(1951–1980) for each of the four monthly-native snow variables.
The headline numbers (summer snow water equivalent collapse, spring snowmelt rise)
are in the **summer** and **spring** rows.

```{r cd-snow-seasonal-table}
snow_monthly <- c("swe", "snowfall", "snowmelt", "snow_cover")
season_order <- c("winter", "spring", "summer", "fall")

snow_seasonal <- cmp_pct[cmp_pct$variable %in% snow_monthly &
                          cmp_pct$period %in% season_order, ]
snow_seasonal$period   <- factor(snow_seasonal$period, levels = season_order)
snow_seasonal$variable <- factor(snow_seasonal$variable, levels = snow_monthly,
                                 labels = c("Snow water equivalent (mm)", "Snowfall (mm)",
                                            "Snowmelt (mm)", "Snow cover (%)"))
snow_seasonal <- snow_seasonal[order(snow_seasonal$variable, snow_seasonal$period), ]
snow_seasonal$mean_a     <- round(snow_seasonal$mean_a, 1)
snow_seasonal$mean_b     <- round(snow_seasonal$mean_b, 1)
snow_seasonal$difference <- round(snow_seasonal$difference, 1)
snow_seasonal <- snow_seasonal[, c("variable", "period", "mean_b",
                                    "mean_a", "difference")]
names(snow_seasonal) <- c("Variable", "Season",
                          "Pre-warming (1951–1980)",
                          "Recent (2015–2025)", "Δ %")

kableExtra::kable_styling(
  knitr::kable(snow_seasonal, label = NA, row.names = FALSE,
    caption = "Seasonal snowpack: recent decade versus pre-warming reference, FWCP Peace Region. Summer snow water equivalent collapse (-75 %) and spring snowmelt rise (+37 %) carry the headline signals."),
  bootstrap_options = c("striped", "hover", "condensed")
)
```

### Annual snowpack signals {-}

Four derived annual scalars capture the snowpack signals the
literature treats as headline metrics for snow hydrology: peak
snowpack [@mote_etal2018Dramaticdeclines;
@mote_etal2005DECLININGMOUNTAIN], the date of melt midpoint
[@stewart_etal2005ChangesEarlier; @cayan_etal2001ChangesOnset],
freshet flashiness, and the fraction of precipitation falling as
snow [@knowles_etal2006TrendsSnowfall].

```{r cd-snow-swe-max, fig.cap = "Annual peak snow water equivalent (swe_max) for the FWCP Peace Region. ERA5-Land mm snow water equivalent, regional spatial mean."}
trn_swe_max <- cd::cd_trend(
  ano[ano$variable == "swe_max" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd::cd_plot_timeseries(ano, variable = "swe_max", period = "annual",
                       trend = trn_swe_max,
                       title = "Annual peak snow water equivalent — Anomaly")
```

```{r cd-snow-doy-50, fig.cap = "Day of year when half the annual snowmelt has accumulated (snowmelt_doy_50). Earlier dates indicate an earlier freshet centroid — directly relevant to crossing-design hydrology and to migration timing."}
trn_doy <- cd::cd_trend(
  ano[ano$variable == "snowmelt_doy_50" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd::cd_plot_timeseries(ano, variable = "snowmelt_doy_50", period = "annual",
                       trend = trn_doy,
                       title = "Snowmelt 50 % DOY — Anomaly")
```

```{r cd-snow-rate-peak, fig.cap = "Annual maximum of 7-day rolling daily snowmelt (snowmelt_rate_peak). Higher values indicate more concentrated freshet pulses on the snowpack side, upstream of routing through soil and channel storage."}
trn_rate <- cd::cd_trend(
  ano[ano$variable == "snowmelt_rate_peak" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd::cd_plot_timeseries(ano, variable = "snowmelt_rate_peak", period = "annual",
                       trend = trn_rate,
                       title = "Peak weekly melt rate — Anomaly")
```

```{r cd-snow-fraction, fig.cap = "Annual snowfall fraction: the percent of annual precipitation that fell as snow. Anomaly is in percentage points relative to the 1951-1980 baseline."}
trn_frac <- cd::cd_trend(
  ano[ano$variable == "snowfall_fraction" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd::cd_plot_timeseries(ano, variable = "snowfall_fraction", period = "annual",
                       trend = trn_frac,
                       title = "Snowfall fraction — Anomaly")
```

### What this means for the FWCP Peace Region {-}

**Snow is leaving the region earlier each year, not falling
less.** Total annual snowfall is roughly stable (-6 %); the
snowpack decline is dominated by *earlier* melt, not by *less*
snow. The seasonal data make this concrete: spring snowmelt is up
37 % while summer snowmelt is down 61 % — same total melt,
redistributed earlier into the year. This matches the broader
Pacific-Northwest signal documented since
@mote_etal2005DECLININGMOUNTAIN.

**The freshet is shifting into spring.** Spring snowmelt up 37 %
is direct evidence of an earlier freshet centroid — consistent in
direction and rough magnitude with the 1–4 week earlier streamflow
timing across western North America in
@stewart_etal2005ChangesEarlier, and with the 10-day Fraser
freshet advance in @kang_etal2016ImpactsRapidly for a basin whose
southern boundary is just south of the Peace.

**Summer is becoming snow-free.** Summer snow water equivalent has collapsed by 75 %
and summer snowmelt is down 61 %. The high-elevation snowpack that
historically lingered into summer no longer does in the recent
decade. For aquatic ecosystems downstream this is a loss of
late-season cold-water input to streams during the warmest part of
the year — the weeks where stream-temperature stress is highest
for cold-water salmonids.

## Spatial pattern of warming {-}

The regional zonal mean reduces a region this size to a single
number; the spatial pattern carries the rest of the story.

Warming is not spatially uniform. Total departures range from
about +1.5 °C in the southeast to over +2.1 °C in the west and
centre (Figure \@ref(fig:cd-map-tmean)). The dominant gradient
runs east-to-west — the western half of the region averages
roughly 0.2 °C more warming than the eastern half — with a smaller
south-to-north component. The west-of-Rockies pattern is
consistent with windward-slope amplification along the Continental
Divide [@pepin_etal2015Elevationdependentwarming].

```{r cd-map-tmean, fig.cap = "Spatial pattern of annual mean temperature departure across the FWCP Peace Region (2015-2025 mean minus 1951-1980 mean, degrees C). Warming is broad but uneven — the west-of-Rockies pattern is consistent with windward-slope amplification along the Continental Divide.", fig.height = 6.5}
ggplot() +
  geom_spatraster(data = departure) +
  geom_sf(data = ecoregions, fill = NA, color = "grey25",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = aoi,   fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(data = lakes, fill = NA, color = "grey40", linewidth = 0.2) +
  geom_sf(data = highways, color = "#9c9c9c", linewidth = 0.6) +
  geom_sf(data = highways, color = "#ffe585", linewidth = 0.4) +
  geom_sf(data = towns, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3, fill = "white", alpha = 0.8
  ) +
  scale_fill_distiller(palette = "RdBu", direction = -1,
                       name = expression(Delta * degree * C)) +
  coord_sf(xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
           ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])) +
  labs(title = "Temperature Departure (2015-2025 vs 1951-1980)") +
  theme_minimal(base_size = 11) +
  theme(axis.title = element_blank())
```

## Per-ecoregion variation {-}

The regional zonal mean averages over five ecoregions with
different elevations and exposures. Running the same pipeline on
each ecoregion polygon tests whether the regional story holds
within each ecoregion — and where it does not.

All five ecoregions warmed at roughly the same rate. Precipitation
is the variable that diverges: the two northernmost ecoregions
(Boreal Mountains and Plateaus, Northern Canadian Rocky Mountains)
show statistically significant precipitation increases, while the
southern and central ecoregions do not. Vapour pressure deficit is
up significantly in all five.

```{r cd-ecoregion-reshape, include = FALSE}
results <- lapply(cd_data$ecoregion, function(x) {
  list(ano = x$ano, trn = x$trn, cmp = x$cmp_pct)
})
results <- results[ecoregions$code]

er_labels <- setNames(
  paste0(ecoregions$code, ": ",
         tools::toTitleCase(tolower(ecoregions$name))),
  ecoregions$code
)
ano_all <- do.call(rbind, lapply(seq_along(results), function(i) {
  o <- results[[i]]$ano
  o$ecoregion <- ecoregions$code[i]
  o
}))
ano_all$ecoregion <- factor(ano_all$ecoregion, levels = ecoregions$code)

build_segments <- function(var) {
  do.call(rbind, lapply(seq_along(results), function(i) {
    sub <- results[[i]]$trn
    sub <- sub[sub$variable == var & sub$period == "annual", ]
    if (nrow(sub) == 0) return(NULL)
    yrs <- ano_all$year[ano_all$variable == var
                        & ano_all$ecoregion == ecoregions$code[i]]
    data.frame(
      ecoregion = factor(ecoregions$code[i], levels = ecoregions$code),
      window    = paste0(sub$trend_start, "–", max(yrs)),
      x_start   = sub$trend_start,
      x_end     = max(yrs),
      y_start   = sub$intercept + sub$slope * sub$trend_start,
      y_end     = sub$intercept + sub$slope * max(yrs)
    )
  }))
}
trn_segs_tmean   <- build_segments("tmean")
trn_segs_prcp    <- build_segments("prcp")
trn_segs_swe_max <- build_segments("swe_max")
trn_segs_doy_50  <- build_segments("snowmelt_doy_50")
```

```{r cd-facet-tmean, fig.cap = "Annual mean temperature anomaly relative to the 1951-1980 baseline, by ecoregion. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present). A solid line steeper than the dashed line indicates accelerating warming.", fig.height = 6}
ggplot(ano_all[ano_all$variable == "tmean" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_tmean,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = expression("Mean temperature anomaly (" * degree * C * ")")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 7))
```

```{r cd-facet-prcp, fig.cap = "Annual precipitation anomaly (% of the 1951-1980 baseline) by ecoregion. The two northernmost ecoregions (BMP, NRM) show statistically significant precipitation increases over the full record; the others do not.", fig.height = 6}
ggplot(ano_all[ano_all$variable == "prcp" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_prcp,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Precipitation anomaly (% of baseline)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 7))
```

### Snow by ecoregion {-}

Two snow metrics are worth viewing per-ecoregion: peak snow water equivalent (annual
snowpack magnitude) and DOY-50 (the day-of-year by which half the
year's snowmelt has already happened — earlier values mean an
earlier freshet). Pair them with the Watershed Groups Across
Ecoregions section below to read each watershed group's dominant
ecoregion signal.

```{r cd-facet-swe-max, fig.cap = "Annual peak snow water equivalent anomaly by ecoregion, relative to the 1951-1980 baseline. Bars are mm of water-equivalent snowpack departure from baseline.", fig.height = 6}
ggplot(ano_all[ano_all$variable == "swe_max" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_swe_max,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Peak snow water equivalent anomaly (mm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 7))
```

```{r cd-facet-doy-50, fig.cap = "Snowmelt 50 % day-of-year (DOY-50) anomaly by ecoregion, relative to the 1951-1980 baseline. Negative values (red) mean the freshet midpoint shifted earlier in the year.", fig.height = 6}
ggplot(ano_all[ano_all$variable == "snowmelt_doy_50" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_doy_50,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "DOY-50 anomaly (days; negative = earlier melt)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 7))
```

```{r cd-rollup}
get_75 <- function(trn, var) {
  trn[trn$variable == var & trn$period == "annual" & trn$trend_start == 1951, ]
}
rollup <- do.call(rbind, lapply(seq_along(results), function(i) {
  res   <- results[[i]]
  cmp_i <- res$cmp
  tmean <- get_75(res$trn, "tmean")
  tmax  <- get_75(res$trn, "tmax")
  tmin  <- get_75(res$trn, "tmin")
  prcp  <- get_75(res$trn, "prcp")
  vpd   <- get_75(res$trn, "vpd")
  data.frame(
    Ecoregion              = ecoregions$code[i],
    `tmean °C/dec`         = round(10 * tmean$slope, 2),
    `tmax °C/dec`          = round(10 * tmax$slope,  2),
    `tmin °C/dec`          = round(10 * tmin$slope,  2),
    `prcp mm/yr`           = round(prcp$slope, 3),
    `prcp p`               = round(prcp$mk_pvalue, 3),
    `vpd hPa/dec`          = round(10 * vpd$slope, 3),
    `vpd p`                = round(vpd$mk_pvalue, 3),
    `prcp % change`        = round(
      cmp_i$difference[cmp_i$variable == "prcp" & cmp_i$period == "annual"], 1),
    `soil moisture % chg`  = round(
      cmp_i$difference[cmp_i$variable == "soil_moisture" & cmp_i$period == "annual"], 1),
    check.names = FALSE
  )
}))
knitr::kable(rollup, row.names = FALSE,
  caption = "Per-ecoregion roll-up over the 75-year window (1951-present): annual mean / daytime maximum / overnight minimum temperature trends (°C per decade); annual precipitation trend (mm per year) and Mann-Kendall p-value; annual vapour pressure deficit trend (hPa per decade) and p-value; and recent (2015-2025) versus pre-warming (1951-1980) percent change for precipitation and soil moisture. All temperature p-values are below 0.001. A tmin slope greater than the tmax slope indicates the day-night asymmetry.")
```

Ecoregion codes: `r paste0(ecoregions$code, " — ", ecoregions$name_tc, collapse = "; ")`.

## Watershed groups across ecoregions {-}

The FWCP Peace Region is reported and managed at the watershed-
group scale — the British Columbia Freshwater Atlas hydrological
reporting unit, and the unit the FWCP funds work in. Sixteen
watershed groups make up the canonical FWCP Peace list. They span
the five ecoregions in different ways: some sit entirely within
one, others split across two or three (Figure
\@ref(fig:cd-map-wsgs)). The table that follows gives, for each
watershed group, the share of its area falling in each ecoregion —
the cross-reference that maps the regional and ecoregion climate
signals onto the per-watershed prioritisation unit.

```{r cd-map-wsgs, fig.cap = "Watershed groups intersecting the FWCP Peace Region (full extent, including parts spilling outside the FWCP boundary), labelled with their codes, on top of ecoregion fills.", fig.height = 6.5}
wsg_centroids <- suppressWarnings(sf::st_centroid(wsgs))
ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi,  fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = wsgs, fill = NA, color = "grey25", linewidth = 0.4) +
  ggrepel::geom_label_repel(
    data = wsg_centroids,
    aes(label = code, geometry = geom),
    stat = "sf_coordinates",
    size = 2.6, fill = "white", alpha = 0.85,
    label.padding = unit(0.15, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
           ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)) +
  labs(title = "Watershed Groups on Ecoregion Backdrop") +
  theme_minimal(base_size = 11) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text  = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

```{r cd-wsg-xref-table}
wsg_xref <- read.csv("data/gis/cd_peace_wsg_ecoregion.csv", check.names = FALSE)
wsg_xref$commentary <- NULL
names(wsg_xref) <- c("Code", "Name", "km² in FWCP",
                     "FAB %", "CRM %", "OMM %", "BMP %", "NRM %")
kableExtra::kable_styling(
  knitr::kable(wsg_xref, label = NA, row.names = FALSE,
    caption = "Watershed groups in the FWCP Peace Region with the percent of each group's area falling in each of the five ecoregions. Rows where one ecoregion is at or near 100 % indicate watershed groups contained within a single climate-physiography zone; rows split across two or more ecoregions indicate watershed groups whose climate departure can pull from more than one regional pattern."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "420px")
```

<br>

## Interpretation for fish passage {-}

Four findings tie the regional climate-departure analysis to
per-watershed barrier prioritisation.

**Warming is broad and uniform; the gradient is small.** All five
ecoregions warmed +1.7 to +1.9 °C since 1951 — temperature is not
the variable that *separates* watershed groups within the Peace.
What it *does* set is the regional context: every reach in the
project area is operating in air ~1.8 °C warmer than the conditions
the channel, floodplain, and fish community evolved under. The
within-region thermal differentiation comes from elevation,
aspect, riparian cover, and groundwater contribution — not from
climate-departure magnitude.

**Precipitation is increasing significantly only in the two
northernmost ecoregions.** Watershed groups dominated by Boreal
Mountains and Plateaus (Toodoggone, Firesteel, Ingenika) or
Northern Canadian Rocky Mountains (Finlay, Fox) sit on
statistically defensible positive precipitation trends. The other
eleven watershed groups sit in Fraser Basin, Central Canadian
Rockies, or Omineca Mountains, where precipitation has not changed
in a way distinguishable from natural variability. For barrier
prioritisation in those eleven, the climate-driven hydrology story
is dominated by *atmospheric drying* (rising VPD, flat soil
moisture, snowpack timing) rather than precipitation magnitude.

**The atmosphere is drying despite, in places, more
precipitation.** Vapour pressure deficit rose significantly in
every ecoregion. Soil moisture is flat across the region even
where precipitation rose — the warmer atmosphere is pulling the
extra water back through evaporation and transpiration. The
ecological implication is that streamflow during the summer
low-flow period — the thermal-stress weeks where cold-water rearing
habitat is at the margin — is not getting the relief that the
precipitation increase, where it occurred, would otherwise suggest.

**Snow is leaving the region earlier, not falling less.** The
snowpack-timing signal is regionally uniform — DOY-50 shifted
earlier in every ecoregion. The freshet — the dominant high-flow
event that mobilises spawning gravels, refills off-channel rearing
habitat, and that crossing structures are sized against — is
arriving weeks earlier across all 16 watershed groups. Summer snow water equivalent
collapsed 75 %; the cold-water input that high-elevation snowpack
provides to streams during the warmest weeks of summer is dropping
in parallel. Both signals act on barrier prioritisation through
hydrology and stream temperature rather than through which
watershed they hit hardest.

For the cold-water resident salmonids the FWCP Peace supports —
bull trout, Arctic grayling, mountain whitefish, rainbow trout,
kokanee — the implication for barrier prioritisation is that two
fundamentally different climate-departure stories play out across
the 16 watershed groups, and that the priority weight on barrier
removal should respond to both: warmer headwater reaches that gain
growing-degree-days have *more* habitat to recover above the
barrier, while reaches sitting closer to the upper thermal niche
have less. The watershed-group ecoregion mapping in the table above
is the bridge between the regional climate-departure summary in the
main body of this report and the per-watershed weight applied to
each barrier.
