---
output: html_document
editor_options:
  chunk_output_type: console
---


# Appendix - Collaborative GIS Layers {-#app-gis}

Background spatial layers and tables loaded to the collaborative GIS project (`r params$gis_project_name`) at the time of writing are listed `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("online within the Collaborative GIS Layers appendix of this report found ", ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "app-gis.html", anchor_text = "here")))}else{knitr::asis_output("in Table \\@ref(tab:tab-rfp-tracking)")}`. The project serves as the spatial integration layer for fieldwork — viewable in QGIS in the office and synced to mobile devices for navigation and standardized data collection in the field via Mergin Maps.

```{r rfp-metadata, eval = params$update_gis}
# grab the metadata
md <- rfp::rfp_meta_bcd_xref()

# burn locally so we don't nee to wait for it
md |>
  readr::write_csv("data/rfp_metadata.csv")
```

```{r rfp-tracking-copy, eval = params$update_gis}

md_raw <- readr::read_csv("data/rfp_metadata.csv")

md <- dplyr::bind_rows(
  md_raw,
  rfp::rfp_xref_layers_custom
)

# first we will copy the doc from the Q project to this repo - the location of the Q project is outside of the repo!!
q_path_stub <- "~/Projects/gis/sern_peace_fwcp_2023/"

# this is differnet than Neexdzii Kwa as it lists layers vs tracking file (tracking file is newer than this project).
# could revert really easily to the tracking file if we wanted to.
gis_layers_ls <- sf::st_layers(paste0(q_path_stub, "background_layers.gpkg"))

gis_layers <- tibble::tibble(content = gis_layers_ls[["name"]])

# remove the `_vw` from the end of content
rfp_tracking_prep <- dplyr::left_join(
  gis_layers |>
    dplyr::distinct(content, .keep_all = FALSE),

  md |>
    dplyr::select(content = object_name, url = url_browser, description),

  by = "content"
) |>
  dplyr::arrange(content)

rfp_tracking_prep |>
  readr::write_csv("data/rfp_tracking_prep.csv")

```

```{r tab-rfp-tracking, eval= gitbook_on}
rfp_tracking_prep <- readr::read_csv(
  "data/rfp_tracking_prep.csv"
)

rfp_tracking_prep |>
  fpr::fpr_kable(caption_text = "Layers loaded to collaborative GIS project.",
                 footnote_text = "Metadata information for bcfishpass and bcfishobs layers can be provided here in the future but currently can usually be sourced from https://smnorris.github.io/bcfishpass/06_data_dictionary.html .",
                 scroll = gitbook_on)
```
